---
title: "Explaining population coverage bias"
vignette: >
  %\VignetteIndexEntry{Explaining population coverage bias}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
bibliography: references.bib
---

## Aim

The [previous vignette](v04-measuring-coverage-bias.html) measured population
coverage bias and showed that some spatial units are better covered by
the digital trace data than others. The next natural question to ask is
whether those differences in coverage are related to the characteristics
of the local populations. The aim in this vignette is to address that
question. To this end, we use a simple regression model as a means to
illustrate the workflow for package users. The research papers
[@cabrera2025; @cabrera2026] associated with the DEBIAS methodology for
bias assessment use more flexible machine-learning methods.

## Load data

Like in the rest of the `debiasR` documentation, we use UK data at the local authority district (LAD) level for illustration. We start by loading the required packages:

```{r setup}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

We then load the necessary datasets from `debiasRdata`: the coverage and the covariates datasets. The coverage dataset contains the benchmark Census population and the observed digital trace user count for each LAD. It stores the LAD identifier in a column called `code`. We rename it to `area` so that it uses the same identifier name as the covariate table and the rest of the `debiasR` workflow.

```{r}
coverage <- debiasRdata::coverage_lad |>
  rename(area = code)

head(coverage)
```

The covariate table contains selected area-level characteristics from the 2021 Census, including percentage of UK-born residents, percentage of residents aged 20-29, percentage of residents aged 70 or over, percentage of residents with Level 4 qualifications, percentage of households without central heating, percentage of residents in routine occupations and percentage of rural land.

```{r}
covariates <- debiasRdata::lad_covariates

head(covariates)
```

Next, we use the `measure_bias()` function to calculate the active-user coverage rate and coverage bias.

```{r}
bias_lad <- measure_bias(coverage)

bias_lad |>
  select(area, population, user_count, coverage_score, coverage_bias) |>
  head()
```

The [previous vignette](v04-measuring-coverage-bias.html) introduced the idea of a proportional baseline. If every LAD were covered at the same overall rate, the local coverage
rate would be the same as the national average. Using the function `validate_bias_residual_structure()`, we can extract the `area_level` object and inspect which locations (LADs) depart from the simple assumption of the proportional baseline model. Positive values of `residual_bias` identify LADs with more coverage bias
than expected under the proportional baseline. Negative values identify
LADs with less coverage bias than expected.

```{r}
bias_residuals <- validate_bias_residual_structure(
  coverage_df = coverage,
  coverage_area_col = "area",
  population_col = "population",
  user_count_col = "user_count"
)

residual_lad <- bias_residuals$area_level |>
  mutate(
    baseline_coverage_bias = 1 - global_coverage_score,
    residual_bias = coverage_bias - baseline_coverage_bias
  ) |>
  left_join(
    coverage |> select(area, name),
    by = "area"
  )

residual_lad |>
  select(area, name, coverage_score, coverage_bias, residual_bias) |>
  head()
```

Next, we can join these residuals to the area-level covariates via a `left_join` that uses the `area` identifier as the key to merge the two relevant datasets.

```{r}
bias_covariates <- residual_lad |>
  left_join(
    covariates |>
      select(
        area,
        per_age_20.29,
        per_age_70plus,
        per_level4,
        per_hh_no_centralheat,
        per_NS_SeC_L13_routine,
        rural_pct
      ),
    by = "area"
  )

bias_covariates |>
  select(
    area,
    residual_bias,
    per_age_20.29,
    per_age_70plus,
    per_level4,
    rural_pct
  ) |>
  head()
```

## Explore one relationship at a time

Before fitting a more complex multivariate model, it is useful to look at simple bivariate relationships. This inspection enables us to see whether residual bias tends to be higher or lower in areas with different local characteristics. For example, below we plot the residual coverage bias against the percentage of population aged between 20 and 29 (left panel) and the percentage of population with level 4 qualifications. While these plots do not prove causality, they help us identify whether coverage bias displays any
meaningful patterns rather than being randomly distributed across
places.

```{r}
#| echo: false
#| fig-width: 10
#| fig-height: 4.5

plot_young_adults <- ggplot(
  bias_covariates,
  aes(x = per_age_20.29, y = residual_bias)
) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Population aged 20-29 (%)",
    y = "Residual coverage bias",
    title = "Young-adult population share"
  ) +
  theme_minimal()

plot_qualifications <- ggplot(
  bias_covariates,
  aes(x = per_level4, y = residual_bias)
) +
  geom_hline(yintercept = 0, colour = "grey50") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Population with level 4 qualifications (%)",
    y = "Residual coverage bias",
    title = "Level 4 qualification share"
  ) +
  theme_minimal()

grid::grid.newpage()
grid::pushViewport(grid::viewport(layout = grid::grid.layout(1, 2)))
print(
  plot_young_adults,
  vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 1)
)
print(
  plot_qualifications,
  vp = grid::viewport(layout.pos.row = 1, layout.pos.col = 2)
)
```



## Fit a simple explanatory model

We now fit a multivariate linear regression. The outcome variable in the regression is `residual_bias`, which compares the coverage bias of an LAD has compared with the proportional baseline. The predictors in this regression are the variables in the covariates dataset:
-   `per_age_20.29`: share of the population aged 20-29
-   `per_age_70plus`: share of the population aged 70 or over
-   `per_level4`: share of the population with Level 4 qualifications
-   `per_hh_no_centralheat`: share of households without central heating
-   `per_NS_SeC_L13_routine`: share of the population in routine
    occupations
-   `rural_pct`: rural population share

Below we fit the model and extract a table of fitted regression coefficients. The sign of each coefficient (estimate) gives the direction of the
association, holding the other variables in the model constant. A
positive coefficient means that higher values of the covariate are
associated with more residual coverage bias (less representation than
average in the mobile phone dataset). A negative coefficient means that
higher values are associated with less residual coverage bias.

```{r}

bias_model <- lm(
  residual_bias ~
    per_age_20.29 +
    per_age_70plus +
    per_level4 +
    per_hh_no_centralheat +
    per_NS_SeC_L13_routine +
    rural_pct,
  data = bias_covariates
)

coefficient_table <- as.data.frame(coef(summary(bias_model)))
coefficient_table$term <- rownames(coefficient_table)
rownames(coefficient_table) <- NULL
names(coefficient_table) <- c("estimate", "std_error", "statistic", "p_value", "term")

coefficient_table |>
  select(term, estimate, std_error, statistic, p_value)
```



## Compare observed and fitted residuals

The fitted values from the model summarise the part of residual bias
that can be described by the selected covariates. But how well is the observed residual bias described by the selected covariates? To answer this question we plot observed vs fitted residual bias. The largest unexplained residuals are areas where the simple model does
not capture the local pattern well. These areas may be useful to inspect
before choosing a bias adjustment strategy.

```{r}
bias_covariates <- bias_covariates |>
  mutate(
    fitted_residual_bias = fitted(bias_model),
    unexplained_residual = residual_bias - fitted_residual_bias
  )

ggplot(bias_covariates, aes(x = fitted_residual_bias, y = residual_bias)) +
  geom_hline(yintercept = 0, colour = "grey80") +
  geom_vline(xintercept = 0, colour = "grey80") +
  geom_point(alpha = 0.7, colour = "#356859") +
  geom_smooth(method = "lm", se = FALSE, colour = "#8B3A3A") +
  labs(
    x = "Fitted residual bias",
    y = "Observed residual bias",
    title =
  ) +
  theme_minimal()
```


```{r echo=FALSE}
# bias_covariates |>
#   arrange(desc(abs(unexplained_residual))) |>
#   select(
#     area,
#     residual_bias,
#     fitted_residual_bias,
#     unexplained_residual
#   ) |>
#   head(10)
```

## Interpretation

This vignette uses a simple regression to "explain" bias. If residual
bias shows a non-trivial relationship with area-level characteristics,
then the variation in bias across geographies is not simply random
noise. Instead, this is a sign that some types of places are more or
less visible in the digital trace data than expected under a
constant-coverage baseline. Therefore, bias adjustment methods need to allow bias to vary across places too.

The regression used here is deliberately simple and descriptive. The
associated research papers [@cabrera2025; @cabrera2026] use more flexible explainable
machine-learning methods to capture nonlinear and interacting
relationships between bias and area-level covariates. The purpose of this vignette is to introduce the same
logic in a way that is easy to inspect, reproduce and adapt.
