---
title: "Adjusting biases in origin-destination flows"
vignette: >
  %\VignetteIndexEntry{Adjusting biases in origin-destination flows}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

## Aim

This vignette describes the set of six adjustment methods included in
`debiasR`, to correct mobile-phone-derived origin-destination mobility
counts. `debiasR` implements five existing methods and a novel Bayesian
multilevel modelling approach. The Bayesian approach comprises a set of
functional form and model options which we explain in this and next
vignette.

This vignette describes how each adjustment method works, associated
functions and parameters, and how to implement it. The vignette focuses
on how to implement the various functions using an small sample of the
dataset introduced in the [Getting set up](v03-getting-set-up.html).
Since we are using a small sample of the dataset, the results in this
vignette do not reflect the real performance of the methods. The next
[Validation](v07-validation.html) vignette illustrates the comparative
performance of the various methods. All the adjustment methods in
`debiasR` use mobile-phone-derived aggregated flows as input data and
return adjusted flows `flow_adj`.

The adjustment methods introduced here differ in fundamental ways, in
terms of data input requirements, how they derive adjustment weights and
the extent they can adjust biases in the data. These differences are
described in more detail below.

::: key-point-box
While most existing flow adjustment approaches are trained on a target
benchmark set of flows, the DEBIAS Bayesian approach does not impose
this constraint. This is its biggest advantage, as the lack of timely
and granular benchmark flow data from traditional data systems is
normally the primary reason to use flows from digital sources.
:::

Hence, methods relying on target benchmark data are of limited use in
live scenarios of emergency, humanitarian or anticipatory action
response. But ultimately, as discussed below, the practical choice of
which methods to use should be guided by what data are available and
validation against a credible benchmark.

## Intuition

Let us quickly recall the challenge we face when working with
mobile-phone-derived mobility flows. These flows typically capture movements associated with a small sample of the resident population, which may not be representative of the underlying population. The idea of the adjustment methods included in `debiasR`
is to adjust the volume of the observed flows to make them more representative of
the benchmark flows, which we consider as the ground truth. Formally, let $F^{mpd}_{ij}$ denote the
observed mobile-phone-derived flow from origin $i$ to destination $j$,
and let $F^{adj}_{ij}$ denote the adjusted flow. If the
observed MPD flow was under-counted with respect to the benchmark flow, a good adjustment methods, would increase $F^{adj}_{ij}$ with respect to $F^{mpd}_{ij}$, to make it more similar to the benchmark flow. If the observed MPD flow was over-counted, the adjustment method would reduce $F^{adj}_{ij}$.

Let us recall the notation:

| Symbol           | Description                                             |
|:-----------------------------------|:-----------------------------------|
| $F^{mpd}_{ij}$   | Observed MPD flow from origin $i$ to destination $j$.   |
| $F^{bench}_{ij}$ | Benchmark flow from origin $i$ to destination $j$.      |
| $F^{adj}_{ij}$   | Adjusted flow returned by an adjustment method.         |
| $P_i, P_j$       | Population at origin and destination.                   |
| $U_i, U_j$       | Active-user count at origin and destination.            |
| $c_i = U_i/P_i$  | Active-user coverage rate.                              |
| $b_i = 1 - c_i$  | Active-user coverage bias.                              |
| $d_{ij}$         | Distance between origin $i$ and destination $j$.        |
| $X_i, X_j$       | Origin and destination covariates, such as rural share. |

## Adjustment methods in `debiasR`

This vignette covers the six bias adjustment methods currently available
in `debiasR`. The table below illustrates how they map onto `debiasR`
functions and relevant references that have inspire our thinking. Below
we explain how each of these methods adjusts biases and how you can
implement them via `debiasR`.

| Method | Function | Reference |
|:-----------------------|:-----------------------|:-----------------------|
| Inverse Penetration Rate Weights | `adjust_inverse_penetration()` | Chi et al. (2025) |
| Selection Rate | `adjust_selection_rate()` | Chi et al. (2025) |
| Selection Rate II | `adjust_selection_rate2()` | Zagheni and Weber (2012) |
| Raking Ratio (or Iterative proportional fitting) | `adjust_raking_ratio()` | Chi et al. (2025) |
| Coefficient regression modelling | `adjust_coefficient()` | Chi et al. (2025) |
| Bayesian multilevel modelling | `adjust_multilevel_bayes()` | Rowe and Cabrera (in progress) |

## Data

We use the same empirical local authority district (LAD) travel-to-work
example introduced in earlier vignettes. The observed origin-destination
flow data comes from Locomizer-derived mobile phone data and the
benchmark origin-destination flows from the 2021 UK Census. The data are
provided through the `debiasRdata` companion package and described in
more detail in [Getting set up](v03-getting-set-up.html).

We start by loading the necessary packages:

```{r load-packages}
library(debiasR)
library(debiasRdata)
library(dplyr)
```

We then identify the main data objects from `debiasRdata`.

```{r}
OD_travel2work <- debiasRdata::lad_OD_travel2work
census_OD_travel2work <- debiasRdata::census_lad_OD_travel2work
coverage <- debiasRdata::coverage_lad |>
  dplyr::rename(area = code)
covariates <- debiasRdata::lad_covariates
centroids <- debiasRdata::lad_centroids
```

```{r}
head(OD_travel2work)
head(census_OD_travel2work)
head(coverage)
```

For the adjustment examples, we use `debiasR_example_data()` to prepare
a compact, aligned version of these data. It creates the complete-grid
OD structure and distance table used by some adjustment methods below, which keeps
the focus on how the `debiasR` functions work.

```{r load-adjustment-data}
example_data <- debiasR_example_data(
  n_areas = 25,
  complete_grid = TRUE
)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage_adjustment <- example_data$coverage
covariates_adjustment <- example_data$covariates
distance <- example_data$distance
```

We structure the outputs in a standard format so they are easier to
interpret and compare:

-   `flow_mpd_raw`: column reporting raw, unadjusted MPD flows used as input.
-   `flow_mpd_adjusted`: column reporting adjusted MPD-derived estimates returned by the adjustment method.
-   `flow_benchmark`: column reporting observed Census benchmark flows used as validation and as input by some of the adjustment methods.

```{r}
benchmark_flow_lookup <- benchmark_od |>
  dplyr::select(
    origin,
    destination,
    flow_benchmark = flow
  )

show_adjustment_comparison <- function(adjusted_df,
                                       extra_cols = character(),
                                       n = 5) {
  adjusted_df |>
    dplyr::left_join(
      benchmark_flow_lookup,
      by = c("origin", "destination")
    ) |>
    dplyr::select(
      origin,
      destination,
      flow_mpd_raw = flow,
      flow_mpd_adjusted = flow_adj,
      flow_benchmark,
      dplyr::any_of(extra_cols)
    ) |>
    dplyr::slice_head(n = n)
}
```

## Adjustment methods

### Inverse penetration rate

The inverse penetration rate adjusts origin-destination flows by scaling
observed mobile-phone-derived flows according to how well the population
is represented in the mobile-phone data. For each area, we calculate the
active-user coverage rate ($c_i$) as the number of observed active users
($U_i$) divided by the benchmark population ($P_i$). The inverse of this
rate becomes the adjustment weight ($w_i$). Areas with lower coverage
receive larger weights because their flows are assumed to be more
undercounted. For example, if only 5% of residents in an origin area
appear in the mobile-phone data, flows from that origin can be scaled up
by a factor of 20; that is $1 / 0.05 = 20$, so each observed flow is
treated as representing about twenty times as many people in the
benchmark population. The adjustment can use origin coverage,
destination coverage or a combination of both. This makes the method
simple and transparent, but it assumes that coverage bias is mainly
proportional to population coverage and does not directly account for
differences in who is represented among mobile-phone users. Formally:

*Penetration rate and inverse penetration rate weight*:

$$c_i = \frac{U_i}{P_i}, \qquad
w_i = \frac{1}{c_i} = \frac{P_i}{U_i}$$

*Origin-weighted adjustment*:

$$F^{adj}_{ij} = w_i F^{mpd}_{ij}$$

*Destination-weighted adjustment*:

$$F^{adj}_{ij} = w_j F^{mpd}_{ij}$$

*Both-side adjustment*:

$$F^{adj}_{ij} = \sqrt{w_iw_j}F^{mpd}_{ij}$$

#### Implementation

To use `adjust_inverse_penetration()`, we input our observed
mobile-phone-derived origin-destination flows via `mpd_od_df` and
coverage data to `coverage_df` in tabular format. The coverage table
should contain the benchmark population and active-user counts needed to
calculate how strongly each area is underrepresented. The `weight_by`
argument lets you choose where the correction is applied: "origin"
adjusts flows using coverage at the origin, "destination" uses coverage
at the destination and "both" combines both sides. The result is a data
frame that preserves the original observed flow in `flow` and adds
`flow_adj`, the adjusted flow after inverse penetration weighting.

If we implement inverse penetration weighting adjusting by origins only,
we set `weight_by` to "origin" as done below. The resulting table
compares the original observed MPD flow, the
inverse-penetration-adjusted estimate and the Census benchmark flow. The
difference between the adjusted and benchmark columns gives a quick
descriptive view of the adjustment, though we assess it formally in the
validation vignette.

```{r}
adj_inverse_penetration <- adjust_inverse_penetration(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  weight_by = "origin"
)

show_adjustment_comparison(adj_inverse_penetration)
```

This method is a useful first adjustment when the main known source of
bias is uneven population coverage across areas. It is especially
helpful when we have reliable benchmark population counts, active-user
counts and benchmark origin-destination flows. In that setting, inverse
penetration weighting provides a transparent way to rescale observed MPD
flows so that areas with lower mobile-phone coverage contribute more
strongly to the adjusted flow estimates. Its main limitation is that it
treats coverage as the correction mechanism. It assumes that the
observed users in an area are broadly representative of the residents
who are missing from the data. However, it does not adjust for
representativeness bias. It cannot directly account for social,
demographic, behavioural or spatial differences in who appears in the
MPD source. If mobile-phone users are systematically different from
non-users, or if the relationship between coverage and mobility varies
across places, this method may reduce undercounting while still leaving
important selection bias in the adjusted flows.

### Selection rate

Selection rate adjusts origin-destination flows by allowing the
adjustment weight to depend on both population coverage and a covariate
that helps explain who appears in the mobile-phone-derived data. The
intuition is that two areas can have the same active-user coverage rate
but still differ in the kinds of people represented in the MPD source.
The selection-rate method therefore adds an area characteristic, such as
rural share, to the coverage correction. It can also use benchmark
origin-destination flows to calibrate the selection parameter, so that
the adjusted flows are closer to a trusted benchmark when one is
available. This makes the method more flexible than inverse penetration
weighting, but it also makes the result more dependent on the quality of
the chosen covariate and the calibration choice. Formally:

*Origin-side selection weight*:

$$w_i^{(O)}(r_t) =
\frac{1}{I_i c_i^{(O)} + (1 - I_i) r_t}$$

*Destination-side selection weight*:

$$w_j^{(D)}(r_t) =
\frac{1}{I_j c_j^{(D)} + (1 - I_j) r_t}$$

*Adjusted flow*:

$$F^{adj}_{ij} = F^{mpd}_{ij}w_{ij}$$

where $I_i$ and $I_j$ are covariate values; $r_t$ is the selection-rate
parameter; and, $w_{ij}$ is the origin, destination or both-side
correction.

#### Implementation

To use `adjust_selection_rate()`, we again input the observed MPD
origin-destination flows through `mpd_od_df` and the population coverage
table through `coverage_df`. The `covariates_df` and `covariate_col`
arguments tell the function which area-level characteristic to use when
estimating differential selection. In this example, we use `rural_pct`,
the rural land-share covariate from `debiasRdata`. While all other
parameters in `adjust_selection_rate()` have defaults, we recommend
users to specify `benchmark_od_df` for a meaningful selection-rate
adjustment. This data frame corresponds to a benchmark OD table used to
calibrate the selection rate parameter `r` in the equations above.

::: key-point-box
**What about if you do not have benchmark flow?**

`adjust_selection_rate()` has two no-benchmark routes:

(1) you can provide `r_global` directly based on prior evidence,
    substantive knowledge or a sensitivity-analysis design.

(2) if both `benchmark_od_df` and `r_global` are omitted, `debiasR` uses
    a fallback based on the overall active-user coverage rate `U / P`.
    This makes the method executable, but the result should be
    interpreted as an exploratory adjustment rather than a calibrated
    estimate.

*Our recommendation* is to calibrate with a credible `benchmark_od_df`
if available. If no benchmark exists, we recommend treating `r_global`
as a sensitivity parameter. Report the fallback value, test a small set
of plausible alternatives and validate the adjusted flows against any
aggregate or available external evidence.
:::

To calibrate MPD flows, `calibration_aggregate` is set to origin. This
means that MDP flows are calibrated *against* origin-level benchmark
totals. This function tries each value in a grid of `r` values
(`r_grid`), adjusts the MPD flows, sums adjusted and benchmark flows by
origin, and chooses the `r_t` that minimises the absolute difference in
origin totals. `r_grid` is the set of candidate values that
`adjust_selection_rate()` tries when it needs to calibrate the global
selection parameter `r_t`. If `calibration_aggregate` is set to `od`, it
skips that origin aggregation. Instead, for each candidate `r_t`, it
compares each adjusted OD cell directly with the benchmark OD cell (as
described in the note below), and picks the `r_t` with the smallest
total OD-level error. The `weight_by` argument controls if the
correction is applied using origin coverage, destination coverage or
both.

```{r}
adj_selection_rate <- adjust_selection_rate(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  covariates_df = covariates_adjustment,
  covariate_col = "rural_pct",
  weight_by = "origin",
  benchmark_od_df = benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate)

tibble::tibble(r_global = attr(adj_selection_rate,
                               "r_global"))
```

The table below reports the original MPD flow, selection-rate adjusted
estimate and Census benchmark flow. Because this example uses
`weight_by = "origin"`, all flows leaving a given origin are adjusted
using that origin's coverage and selected covariate value. The reported
`r_global` attribute is the calibrated selection-rate parameter chosen
by the function. This method is useful when coverage alone is too simple
and we believe that the data representativeness varies prominently with
an area characteristic. Its main limitation is that the chosen covariate
must be meaningful. If the covariate is weakly related to selection into
the MPD source, the adjustment may add complexity without reducing
coverage or selection bias. The `weight_missing` column in the
`adj_selection_rate` is a diagnostic label. If origin coverage data are
missing and/or invalid, this leads to missing and/or invalid weight
inputs.

### Selection rate II

Selection rate II adjusts origin-destination flows by applying a compact
coverage correction controlled by a single curvature parameter, $k$. The
method keeps the focus on population coverage. Instead of using the
simple inverse of the coverage rate, it transforms the coverage rate
through a nonlinear correction factor. This gives users a middle ground
between inverse penetration weighting and the fuller covariate-based
selection-rate model. When benchmark origin-destination flows are
available, `debiasR` can calibrate $k$ so that the adjusted flows better
match a trusted comparison. Formally:

*Coverage correction factor*:

$$CF(c; k) =
c\frac{\exp(-k) - 1}{\exp(-kc) - 1}$$

where $CF(c; k)$ is the coverage correction factor, $c$ is the
active-user coverage rate for the selected origin or destination area,
and $k$ is the curvature parameter that controls how strongly coverage
is transformed.

*Adjusted flow*:

$$F^{adj}_{ij} = F^{mpd}_{ij}CF(c; k)$$

The same correction can be applied by origin, destination or both,
depending on `weight_by`.

#### Implementation

To use `adjust_selection_rate2()`, the function requires MPD flows and
coverage data as inputs. Input the observed MPD flows to `mpd_od_df` and
coverage data to `coverage_df`. The `weight_by` argument controls
whether the correction uses origin coverage, destination coverage or
both. You can set `k` directly if you have a chosen correction
curvature, or leave it as `NULL` and provide `benchmark_od_df` so the
function calibrates `k` over `k_grid`. As above,
`calibration_aggregate = "origin"` calibrates against origin-level
benchmark totals. The parameter `k` is the analog to `r` in
`adjust_selection_rate()`.

```{r}
adj_selection_rate2 <- adjust_selection_rate2(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  weight_by = "origin",
  benchmark_od_df = benchmark_od,
  calibration_aggregate = "origin"
)

show_adjustment_comparison(adj_selection_rate2)

tibble::tibble(k = attr(adj_selection_rate2, "k"))
```

The output table reports the observed MPD count, adjusted flow and
Census benchmark flow. In this example, the correction applied to a flow
depends on the origin area's coverage rate and calibrated $k$ value. The
reported `k` attribute shows the curvature parameter used by the
adjustment. This method is useful when coverage is the dominant concern
and a compact correction is adequate. Its main limitation is that one
parameter may be too simple when selection bias varies strongly across
places or population groups, and no benchmark data exist.

### Raking ratio

Raking ratio is also known as iterative proportional fitting (IPF).
Raking ratio adjusts origin-destination flows by iteratively rescaling
the observed MPD matrix so that its origin and destination totals match
trusted benchmark margins. The intuition is different from coverage
weighting. Instead of calculating a weight from active-user coverage,
the method asks the adjusted origin-destination matrix to satisfy known
aggregate constraints. If we trust total outflows from each origin and
total inflows to each destination, IPF updates the OD cells until the
row and column totals agree with those benchmarks. Formally:

*Margin constraints*:

$$\sum_j F^{adj}_{ij} = O_i^{bench},
\qquad
\sum_i F^{adj}_{ij} = D_j^{bench}$$

where $O_i^{bench}$ and $D_j^{bench}$ are benchmark outflow and inflow
totals.

#### Implementation

To use `adjust_raking_ratio()`, the function requires MDP flows and set
of constraints as inputs. Feed the observed MPD flows to `mpd_od_df` and
either provide `benchmark_od_df` or supply explicit `origin_targets` and
`destination_targets`. We use `benchmark_od_df`, so `debiasR` derives
the target origin and destination margins from the benchmark OD table.

The `max_iter` and `tol` arguments control the iterative fitting
process. `max_iter` sets the maximum number of update cycles and `tol`
sets the convergence tolerance.

```{r}
adj_raking_ratio <- adjust_raking_ratio(
  mpd_od_df = mpd_od,
  benchmark_od_df = benchmark_od
)

show_adjustment_comparison(adj_raking_ratio)

adj_raking_ratio |>
  dplyr::summarise(
    raw_total = sum(flow, na.rm = TRUE),
    adjusted_total = sum(flow_adj, na.rm = TRUE),
    median_weight = median(weight_ipf, na.rm = TRUE),
    ipf_converged = attr(adj_raking_ratio, "ipf_converged"),
    ipf_iterations = attr(adj_raking_ratio, "ipf_iterations")
  )
```

The output table shows how individual observed MPD flows have been
converted into adjusted flows and how those adjusted flows compare with
the Census benchmark. The summary reports whether the IPF routine
converged, how many iterations it used, and how much the total flow
changed after adjustment. Raking is useful when trusted marginal totals
are available and consistency with those totals is a priority. Its main
limitation is that it requires observed benchmark origin and destination
flow data. A second limitation is that matching origin and destination
margins does not guarantee that the internal allocation between specific
origin-destination pairs is correct, so it should still be validated at
the OD-pair level.

### Coefficient regression modelling

Coefficient regression modelling adjusts origin-destination flows by
learning a direct empirical relationship between observed MPD flows and
benchmark flows. The method is useful when a trusted benchmark OD table
is available for the same origin-destination pairs. Instead of using
population coverage or margins, it fits a regression model that
estimates how MPD counts map onto benchmark counts, then applies that
fitted relationship to the MPD table. In its simplest form, this is a
single multiplicative coefficient. In count-model versions, the
relationship is estimated through a log link. Flow measured as
proportions or probability as, for example, the number of outflows over
the population at risk could be used as input, as well. Formally:

*Core proportional form*:

$$E(F^{bench}_{ij}) = \beta F^{mpd}_{ij}$$

*Count-model form*:

$$\log(\mu_{ij}) = \alpha + \log(F^{mpd}_{ij})$$

so that:

$$\mu_{ij} = e^{\alpha}F^{mpd}_{ij},
\qquad
\beta = e^{\alpha}$$

#### Implementation

To use `adjust_coefficient()`, the essential input requirements are the
MPD and benchmark flows. Input the observed MPD flows to `mpd_od_df` and
benchmark OD flows to `benchmark_od_df`. The `model_family` argument
controls the regression family used to learn the MPD-to-benchmark
relationship. This example uses `"ols"` for a simple proportional
calibration. The `level` argument controls whether the coefficient is
estimated at the OD, origin or destination level. Additional options,
such as `fit_intercept` and `by_source`, let users decide whether to
estimate an intercept or source-specific coefficients.

```{r}
adj_coefficient <- adjust_coefficient(
  mpd_od_df = mpd_od,
  benchmark_od_df = benchmark_od,
  model_family = "ols",
  level = "od"
)

show_adjustment_comparison(adj_coefficient)

attr(adj_coefficient, "coef")
```

The output table shows the original observed MPD flow, regression
adjusted estimate and Census benchmark flow. The coefficient attribute
reports the fitted scaling relationship used to convert MPD flows into
benchmark-like flows. This method is easy to validate because the
benchmark is part of the model-fitting process. Its main limitation is
that it requires benchmark flow data, and it does not adjust for
representativeness bias. A simple coefficient can miss local
heterogeneity unless the chosen model family and estimation level
capture the relevant structure.



### Bayesian multilevel modelling

Bayesian multilevel modelling in `debiasR` is a flexible modelling
framework. In simple terms, the model addresses the following question: given the observed MPD flow and the level of population coverage, what underlying flow is likely to have produced that observation? If sufficient information is available about the characteristics of the locations, the model can also use patterns shared across origins, destinations and OD pairs to improve the estimates.

::: key-point-box
**Important!** The Bayesian models are designed for data-scarce settings where benchmark OD flows are unavailable for adjustment, as often occurs during natural hazards, armed conflict or health emergencies. Unlike all the other adjustment methods introduced in this vignette, the Bayesian method can estimate adjusted flows without benchmark flow data and quantify the uncertainty of those estimates. If benchmark flows are available, they can still be used for external validation.

Therefore, the performance of Bayesian models should not only be assessed on better error performance over benchmark-calibrated methods, but also by considering the provision of uncertainty estimates and flexibility across data sources and time periods.
:::

The models are implemented via the function `adjust_multilevel_bayes()` which enables three model variants and four input data scenarios (S1-S4). Detailed information about model variants and input data scenarios is provided in the [advanced Bayesian adjustment vignette](v09-advanced-bayesian-adjustment.html). The function also supports `model_engine = "frequentist"`, which can help users explore model specifications, covariates, random effects and differences across data sources or time periods before running the more computationally demanding Bayesian models. A practical guide to the `adjust_multilevel_bayes()` variants is provided in the table below:

```{r echo=FALSE}
bayesian_option_guide <- tibble::tribble(
  ~`Model variant`, ~`Use it when`, ~`Intuition`,
  "`observation_model= \"coverage_offset\"`",
  "Estimate real population flows from phone counts, accounting for how well each area is represented in the mobile-phone data.",
  "The model estimates true flows while treating source/time coverage as a fixed observation offset.",
  "`observation_model= \"reduced_form\"`",
  "Adjust the observed phone counts without using coverage rates to estimate how many people those phone counts represent.",
  "The model fits observed MPD flows directly and predicts a counterfactual MPD flow with the coverage bias term set to zero.",
  "`observation_model= \"latent_two_level\"`",
  "Estimate shared population flows when the same OD pairs are observed more than once, such as across different data sources or time periods.",
  "The model estimates source-invariant latent OD or OD-time true-flow states and treats observed source/time rows as noisy measurements.",
  "`model_engine= \"frequentist\"`",
  "Run the frequentist option to check that the data, formulas and assumptions work before running the full Bayesian model.",
  "This option uses the same non-latent input structure for rapid specification checks and method comparison."
)

knitr::kable(bayesian_option_guide)
```


In this section, we focus on illustrating the default model with variant `coverage-offset` and the input data scenario S1, which relies on a single source of MPD data and a single time period.  The coverage-offset model has two components: an **MPD observation model** and a **true-flow model**. The observation model describes how an underlying population flow becomes partially visible in the MPD data. The true-flow model estimates the population mobility pattern behind the observed MPD counts.

***Component 1: MPD observation model.*** The MPD observation model describes how an underlying population flow becomes partially visible in the MPD data:

$$
F^{mpd}_{ij}
\sim
\operatorname{Poisson}
\left(q_{ij}\lambda^{true}_{ij}\right)
$$

Here, $F^{mpd}_{ij}$ is the observed MPD flow from origin $i$ to
destination $j$. The term $\lambda^{true}_{ij}$ is the expected
underlying population flow for the same OD pair. We use
$\lambda^{true}_{ij}$ rather than $F^{true}_{ij}$ because this is the expected rather than observed true flow, which may not be a whole number. The term $q_{ij}$ is the proportion of the underlying flow expected to
be visible in the MPD data. Here, we use the Poisson distribution, as it allows the observed count
to vary around its expected value rather than assuming that it must
equal that value exactly. If the data show greater variation or more
zero flows, `adjust_multilevel_bayes()` also supports negative binomial,
zero-inflated Poisson and zero-inflated negative binomial models. The expected observed MPD flow is therefore: $E[F^{mpd}_{ij}] = q_{ij}\lambda^{true}_{ij}$, or taking logarithms for a linearised form:

$$
\log( E\left[F^{mpd}_{ij}\right])
=
\log(q_{ij})+\eta^{true}_{ij}
$$

where $\lambda^{true}_{ij} = \exp(\eta^{true}_{ij})$. The coverage rate $q_{ij}$ is supplied as known information about the observation process. It tells the model what proportion of the underlying flow the MPD system is expected to observe. Then, for example, if the expected underlying population flow is 500 and $q_{ij}=0.20$, the model expects to observe approximately $0.20\times500=100$ movements in the MPD data. When fitting the model, the reasoning works in the opposite direction, as the observed MPD count and the coverage rate are used to estimate the underlying true population flow that could plausibly have produced that observation.

In `adjust_multilevel_bayes()`, the `coverage_scale` argument determines how $q_{ij}$ is calculated. If `coverage_scale = "origin"`, then $q_{ij}=c_i=U_i/P_i$, where $U_i$ is the observed active-user count and $P_i$ is the benchmark population in the origin area. In this case, all flows leaving the same origin use that origin's coverage rate. Other options are `coverage_scale = "destination"` ($q_{ij} = c_j$) and `coverage_scale = "both"` ($q_{ij} = \sqrt{c_i c_j}$).

***Component 2: True-flow model.*** The second component estimates the expected underlying population flow for each OD pair. We model $\eta^{true}_{ij}$, i.e. the log of the expected flow, as

$$
\eta^{true}_{ij}
=
\alpha
+ \beta_oX_i
+ \beta_dX_j
+ \gamma\log d_{ij}
+ u_i
+ v_j
+ w_{ij}
+ \xi_{ij}
$$

where $\alpha$ represents the overall level of flow, the terms $X_i$ and $X_j$ represent characteristics of the origin and destination and $d_{ij}$ is the distance between them. The terms $u_i$, $v_j$ and $w_{ij}$ give the model a multilevel structure and allow the model to represent patterns shared by flows from the same origin, to the same destination or repeated observations of the same OD pair. When included, flows from related observations can help inform one another. For example, if only a few movements are observed for one OD pair, the model can also consider the general pattern among flows from the same origin or to the same destination. The estimate for that OD pair is therefore based partly on its own data and partly on patterns found in related flows. This can lead to more stable estimates when some observed flows are small or highly variable. After combining all the terms in the above equation, an exponential transformation can be applied to obtain a positive expected population flow, $\lambda^{true}_{ij}$.

***How are the unknown quantities in Component 1 and 2 estimated?***. Following a Bayesian approach, `adjust_multilevel_bayes()` assigns prior distributions to the model parameters, including the regression coefficients and any multilevel effects if present in the model. These priors describe plausible parameter values before the observed MPD flows are considered. The function then combines the prior distributions with the observed MPD flows through the observation model to generate posterior distributions of the parameter values.

$$
\text{prior distributions}
+
\text{observed MPD flows}
\longrightarrow
\text{posterior distributions}
$$

To generate a posterior distribution of the adjusted flow, the function `adjust_multilevel_bayes()` samples many sets of parameter values from their posterior distributions. For each sample, known as posterior draw $s$, a corresponding expected underlying flow is calculated as $F^{adj,(s)}_{ij}$ = $\lambda^{true,(s)}_{ij}$ = $\exp(\eta^{true,(s)}_{ij})$. For example, different posterior draws might produce adjusted flows of 482, 510, 495 and 536 for the same OD pair. These represent different underlying flows that remain plausible given the observed MPD data, coverage rates, model specification and prior assumptions. The collection of draws forms the posterior distribution of the adjusted flow.

#### Implementation

The code below shows how to fit the Bayesian coverage-offset model and inspect its main outputs. When fitting the Bayesian models, it is important to use enough iterations and chains for the application and check the sampler diagnostics before interpreting the results. To use `adjust_multilevel_bayes()`, input the observed MPD flows to `mpd_od_df`, population coverage to `coverage_df`, area-level covariates
to `covariates_df` and OD distances to `distance_df`. The example below
uses the S1 scenario, with a single data source and a single observation period.

For the coverage-offset model variant, most users only need a small set of arguments. The `mobility_formula` defines the true-flow model described above. Here we use
`rural_pct_o`, `rural_pct_d` and `log_distance` to represent origin
covariates, destination covariates and distance. This example uses a
fixed-effect intercept, but the model can accommodate multilevel terms
such as random effects by origin `(1 | origin)` or by destination `(1 | destination)`. The recommended mode for the observation model is `target_scale = "true_flow"` and
`observation_model = "coverage_offset"`. The wider parameter set is covered
in the advanced Bayesian vignette.

```{r echo=FALSE}
bayesian_parameter_guide <- tibble::tribble(
  ~Argument, ~Description,
  "`mobility_formula`", "Defines the true-flow predictor, such as origin covariates, destination covariates, distance and optional multilevel terms.",
  "`target_scale = \"true_flow\"`", "Returns adjusted flows on the estimated true-flow scale rather than on the MPD observation scale.",
  "`observation_model = \"coverage_offset\"`", "Treats MPD flows as coverage-scaled observations of true flows.",
  "`coverage_scale`", "Chooses whether the fixed coverage offset uses origin coverage, destination coverage or both.",
  "`model_engine = \"bayesian\"`", "Fits the posterior model. Use the frequentist option, `model_engine = \"frequentist\"`, first when you want a specification check.",
  "`scenario`, `source_col`, `time_col`", "Declare the source/time data structure. The example below uses S1: one source and one time period.",
  "`prediction_scope`", "Chooses whether predictions are returned only for observed MPD rows or for a supplied complete OD matrix.",
  "`random_intercept`", "Adds or omits origin, destination, OD, source or time pooling structure when the data support it.",
  "`model_family`", "Chooses the count likelihood family, such as Poisson or negative binomial.",
  "`iter`, `chains`, `seed`", "Control Bayesian sampling and reproducibility."
)

knitr::kable(bayesian_parameter_guide)
```

The example below uses a Bayesian fixed-effect true-flow model with an
origin coverage offset.

```{r eval=FALSE}
mpd_s1 <- mpd_od |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

coverage_s1 <- coverage_adjustment |>
  dplyr::mutate(
    mpd_source = "operator_a",
    mpd_time = "2021_q1"
  )

adj_multilevel <- adjust_multilevel_bayes(
  mpd_od_df = mpd_s1,
  coverage_df = coverage_s1,
  covariates_df = covariates_adjustment,
  distance_df = distance,
  mobility_formula = ~ rural_pct_o + rural_pct_d + log_distance,
  bias_formula = ~ bias_e_origin,
  target_scale = "true_flow",
  observation_model = "coverage_offset",
  coverage_scale = "origin",
  model_engine = "bayesian",
  scenario = "s1",
  source_col = "mpd_source",
  time_col = "mpd_time",
  repeated_observation = "none",
  prediction_scope = "complete_grid",
  random_intercept = "none",
  model_family = "poisson",
  flow_adj_summary = "median",
  include_flow_adj_draws = TRUE,
  iter = 1000,
  chains = 2,
  seed = 123,
  refresh = 0
)
```


```{r echo=FALSE}
bayesian_example_file <- system.file(
  "extdata",
  "v06-bayesian-s1-comparison.csv",
  package = "debiasR"
)

if (identical(bayesian_example_file, "")) {
  stop(
    "Missing Bayesian example output."
  )
}

bayesian_example <- tibble::as_tibble(
  utils::read.csv(bayesian_example_file, check.names = FALSE)
)

knitr::kable(bayesian_example, digits = 2)
```

The table above reports the main outputs from
`adjust_multilevel_bayes()`, along with `flow_benchmark` reporting the
benchmark flows for comparison. The column `flow` is the observed MPD count and `flow_adj`
is the main adjusted flow estimate. The function call sets
`flow_adj_summary = "median"`, so `flow_adj` equals `flow_adj_median` in
this example. It is also possible to set `flow_adj_summary = "mean"`. The columns `flow_adj_median` and `flow_adj_mean` report
two posterior summaries of the estimated population flow. The column
`flow_mpd_pred` reports the fitted value on the MPD-observed scale,
after applying the coverage offset. The table also
includes `flow_true_pred`, which in
this example, is equal to `flow_adj`, so the displayed
table keeps `flow_adj` as the single main population-scale adjusted
estimate. Furthemore, the predicted median and mean flow in the MPD data and confidence intervals for the adjusted flows are reported under `flow_mpd_pred_median`, `flow_mpd_pred_mean`, `flow_adj_q2.5`,
`flow_adj_q97.5`, `flow_mpd_pred_q2.5` and `flow_mpd_pred_q97.5`.  The example uses moderate sampler settings for illustrative purposes. For applied analysis, it is recommended to increase the number of iterations and chains.

You can inspect the results metadata to confirm the scenario, model variant, coverage scale and other attributes of model setup.

```{r echo=FALSE}
bayesian_metadata_file <- system.file(
  "extdata",
  "v06-bayesian-s1-metadata.csv",
  package = "debiasR"
)

bayesian_terms_file <- system.file(
  "extdata",
  "v06-bayesian-s1-model-terms.csv",
  package = "debiasR"
)

if (identical(bayesian_metadata_file, "") || identical(bayesian_terms_file, "")) {
  stop(
    "Missing Bayesian example metadata."
  )
}

bayesian_metadata <- tibble::as_tibble(
  utils::read.csv(bayesian_metadata_file, check.names = FALSE)
)

bayesian_terms <- tibble::as_tibble(
  utils::read.csv(bayesian_terms_file, check.names = FALSE)
)

```

```{r eval=FALSE}
attr(adj_multilevel, "result_metadata")
```

```{r echo=FALSE}
knitr::kable(bayesian_metadata)
```

It is also possible to inspect the model terms, which include the specific model equations used for modelling.

```{r eval=FALSE}
attr(adj_multilevel, "model_terms")
```

```{r echo=FALSE}
knitr::kable(bayesian_terms)
```

Finally, inspecting sampler diagnostics can be helpful before interpreting the posterior summaries.

```{r eval=FALSE}
attr(adj_multilevel, "diagnostics")
```

<!-- *More complex Bayesian structures* -->
<!-- The function `adjust_multilevel_bayes()` can accommodate more complex S2-S4 -->
<!-- source/time structures through `scenario`, `source_col`, `time_col` and -->
<!-- `repeated_observation`. For repeated source/time data, users can select -->
<!-- `observation_model = "latent_two_level"`. This model structure estimates one -->
<!-- shared latent true-flow state for an OD pair or OD-time group and treats each -->
<!-- source/time MPD row as an observation of that state. This is most useful for S3 -->
<!-- and S4 structures, where multiple MPD sources observe the same underlying -->
<!-- mobility state. Source and time effects live in the observation model, so -->
<!-- `flow_true_pred` is source-invariant within a latent state, while -->
<!-- `flow_mpd_pred` can vary by source and time. See the [advanced Bayesian -->
<!-- adjustment vignette](v09-advanced-bayesian-adjustment.html) for more -->
<!-- details. -->




## Fit all methods

We can also run all six adjustment methods at once using
`adjust_all_methods()`. The function uses the frequentist option for the
multilevel method by default, which compares the main adjustment methods
without posterior sampling. You can switch `multilevel_engine` to
`"bayesian"` for more complex structures.

```{r}
method_results <- adjust_all_methods(
  mpd_od_df = mpd_od,
  coverage_df = coverage_adjustment,
  benchmark_od_df = benchmark_od,
  covariates_df = covariates_adjustment,
  distance_df = distance,
  covariate_col = "rural_pct",
  multilevel_engine = "frequentist"
)
```

```{r echo=FALSE}
method_engine <- function(x) {
  engine <- attr(x, "model_engine", exact = TRUE)
  if (is.null(engine)) "deterministic" else engine
}

tibble::tibble(
  method = names(method_results),
  engine = vapply(method_results, method_engine, character(1)),
  rows = vapply(method_results, nrow, integer(1)),
  adjusted_total = vapply(
    method_results,
    function(x) sum(x$flow_adj, na.rm = TRUE),
    numeric(1)
  )
)
```

## Method summary

As described above, each adjustment method in `debiasR` has different
assumptions, data needs, strengths and limitations. The table below
focuses on the main differences readers need when choosing a method:
whether the method uses benchmark information during fitting, why the
method is useful, and where it is most likely to fall short.

```{r results='asis', echo=FALSE}
method_summary <- tibble::tribble(
  ~Method, ~`Benchmark role`, ~Rationale, ~Strengths, ~Limitations,
  "Inverse Penetration Rate Weights",
  "Not benchmark-trained. Uses benchmark population and active-user counts, but not benchmark OD flows.",
  "Inflates observed flows using population-to-user ratios where coverage is thin.",
  "Simple, transparent, fast, and does not require benchmark OD flows.",
  "Can overinflate sparse areas and misses compositional selection bias.",

  "Selection Rate",
  "Optionally benchmark-assisted. Can calibrate the selection parameter against benchmark OD flows; otherwise uses coverage and covariates.",
  "Uses coverage, covariates, and optional benchmark calibration to model differential selection.",
  "More flexible than inverse penetration and can use benchmark calibration.",
  "Depends on covariate quality, calibration choices, and a meaningful selection parameter.",

  "Selection Rate II",
  "Optionally benchmark-assisted. Can calibrate the global <code>k</code> parameter against benchmark OD flows; otherwise uses a supplied or default <code>k</code>.",
  "Applies a compact coverage correction with optional <code>k</code> calibration.",
  "Parsimonious, interpretable, and lighter than a covariate-rich adjustment.",
  "Less flexible when bias varies strongly across areas.",

  "Raking Ratio (or Iterative proportional fitting)",
  "Benchmark-margin assisted. Uses benchmark origin and destination margins, or user-supplied targets, rather than fitting to each OD cell.",
  "Uses iterative proportional fitting so adjusted flows match trusted origin and destination margins.",
  "Guarantees consistency with supplied marginal totals.",
  "Matching margins does not guarantee correct destination allocation within each origin.",

  "Coefficient regression modelling",
  "Benchmark-trained. Fits the MPD-to-benchmark relationship using benchmark OD cells.",
  "Learns a direct mapping from MPD flows to benchmark flows and applies it to the MPD table.",
  "Direct benchmark-driven calibration with several model families.",
  "Benchmark quality is critical and a global coefficient can miss local heterogeneity.",

  "Bayesian multilevel modelling",
  "Not benchmark-trained. Uses coverage and covariates during fitting; benchmark OD flows are reserved for external validation.",
  "Fits a count model where active-user coverage enters as a fixed observation-process offset, then returns the estimated true-flow scale.",
  "Richer structure, no benchmark OD needed for fitting, posterior uncertainty, and a path to source/time-aware partial pooling.",
  "Heavier runtime and dependencies; posterior fitting requires sampler diagnostics and should be compared externally against benchmark-calibrated methods."
)

table_html <- knitr::kable(
  method_summary,
  format = "html",
  escape = FALSE,
  table.attr = "class='table method-summary-table'"
)

cat("<div class='method-table-wrap'>", table_html, "</div>")
```

```{=html}
<style>
main table {
  font-size: calc(0.9rem - 1.5px);
  line-height: 1.45;
}

main table th,
main table td {
  vertical-align: top;
}

.key-point-box {
  margin: 1.25rem 0;
  padding: 1rem 1.25rem;
  border: 1px solid #d8dee6;
  border-left: 4px solid #446e9b;
  border-radius: 4px;
  background: #f7f9fb;
}

.key-point-box > :last-child {
  margin-bottom: 0;
}

.method-table-wrap {
  overflow-x: auto;
  margin: 1.25rem 0 1.75rem 0;
  padding: 0.6rem 0.8rem;
  border: 1px solid #e5e7eb;
  border-radius: 8px;
  background: #fff;
}

.method-summary-table {
  min-width: 1050px;
}

.method-summary-table th,
.method-summary-table td {
  padding: 0.75rem;
  white-space: normal;
}

</style>
```
