---
title: "Validation"
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{quarto::html}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
editor:
  markdown:
    wrap: 72
---

```{=html}
<style>
.template-quarto table {
  font-size: 14.5px;
}
.validation-table-small,
.validation-table-small table,
table.validation-table-small,
.template-quarto .validation-table-small {
  font-size: 13px;
}
.validation-table-small table {
  width: 100%;
  table-layout: fixed;
}
.validation-table-small col:nth-child(1) {
  width: 22% !important;
}
.validation-table-small col:nth-child(2) {
  width: 48% !important;
}
.validation-table-small col:nth-child(3) {
  width: 30% !important;
}
.validation-table-small th:nth-child(1),
.validation-table-small td:nth-child(1) {
  width: 22%;
}
.validation-table-small th:nth-child(2),
.validation-table-small td:nth-child(2) {
  width: 48%;
}
.validation-table-small th:nth-child(3),
.validation-table-small td:nth-child(3) {
  width: 30%;
  font-size: 12px;
}
.validation-table-compact,
.validation-table-compact table,
table.validation-table-compact,
.template-quarto .validation-table-compact {
  font-size: 13.5px;
}
.key-point-box {
  margin: 1.25rem 0;
  padding: 1rem 1.25rem;
  border: 1px solid #d8dee6;
  border-left: 4px solid #446e9b;
  border-radius: 4px;
  background: #f7f9fb;
}
.key-point-box > :last-child {
  margin-bottom: 0;
}
</style>
```

## Aim

This vignette describes how to use the `debiasR` functions to assess and
validate adjusted origin-destination (OD) flows derived from mobile
phone data. The validation procedures compare the outcome of the bias
adjustment methods included in `debiasR` against a set of target
benchmark flows (i.e. adjusted *versus* benchmark flows). This tells us
how close the final adjusted output is to a chosen target benchmark.
While the comparison between adjusted *versus* benchmark flows is the
primary focus, `debiasR` also enables two additional comparisons: raw
mobile-phone-derived *versus* benchmark flows, which tells us about the
extent of bias before adjustment; and raw mobile-phone-derived *versus*
adjusted flows, which tells us the extent of adjustment applied to the
original raw data.

::: key-point-box
**Important!** This vignette focuses on illustrating the use of
`debiasR` validation functions. The focus is not on providing an
exhaustive comparison of adjustment methods. Such comparison would
require a more comprehensive setup. So do not take the results provided
here as evidence to prefer one model over another. A key issue to note
is that we are comparing adjustment methods that require benchmark data
as input (so they can "see" the target benchmark flow data) against
Bayesian models, which do not require such input.
:::

## Intuition {#sec-overview-levels}

To validate adjusted flows, the `debiasR` framework is organised based
on five validation levels, moving from *coarse* overall checks based on
summaries across the whole system of flows, to more granular checks at
the local area level. In this way, `debiasR` recognises the complexity
of the validation process, which cannot rely on a single number.
Adjusted flows can look good overall and still perform poorly for some
origins, destinations or individual OD pairs. In particular, we work
with five validation levels.

-   **Level 1** summarises performance across the whole OD system. This
    tells us whether a method is broadly close to the benchmark.
-   **Level 2** compares origin and destination totals, checking whether
    adjusted flows reproduce total outflows from each origin and total
    inflows to each destination.
-   **Level 3** compares individual OD pairs, showing where large local
    errors remain after adjustment.
-   **Level 4** assesses whether each origin allocates its flows across
    destinations in a similar way to the benchmark. This is about the
    shares across destinations, not only about the total scale.
-   **Level 5** checks whether remaining residuals show systematic
    patterns after adjustment. For example, residuals may be larger for
    large benchmark flows, clustered across neighbouring areas, or
    associated with an area covariate.

At each level, it is useful to separate error metrics from correlation
metrics. Error metrics help us understand whether the numbers are close
in magnitude, whereas correlation metrics tell us whether the broader
data pattern is similar. A method can have a high correlation with the
benchmark and still be too high or too low in absolute terms.

## List of validation measures {#sec-list-measures}

Let $F^{adj}_{ij}$ be the adjusted flow from origin $i$ to destination
$j$, and let $F^{bench}_{ij}$ be the benchmark flow for the same OD
pair. Define the error as $e_{ij} = F^{adj}_{ij} - F^{bench}_{ij}$.
Positive errors mean that the adjusted flow is larger than the benchmark
and negative errors that it is smaller. The table uses
adjusted-versus-benchmark notation.[^1]

[^1]: The equations are written for adjusted-versus-benchmark flows. The
    same formulas can be reused for raw-versus-benchmark or
    raw-versus-adjusted comparisons by replacing $F^{adj}_{ij}$ and
    $F^{bench}_{ij}$ with the two flow columns being compared.

::: validation-table-small
| Metric | Description | Equation |
|:-----------------------|:-----------------------|:-----------------------|
| Mean error | Average signed error; shows whether flows are overestimated or underestimated on average. | $\dfrac{1}{n}\sum_{ij} e_{ij}$ |
| Mean absolute error | Average absolute error; shows the typical size of the error, ignoring sign. | $\dfrac{1}{n}\sum_{ij} \lvert e_{ij} \rvert$ |
| Root mean squared error | Error metric that gives more weight to large errors. | $\sqrt{\dfrac{1}{n}\sum_{ij} e_{ij}^{2}}$ |
| Median absolute error | Middle absolute error; less affected by a small number of very large errors. | $\mathrm{median}(\lvert e_{ij} \rvert)$ |
| Mean absolute percentage error | Average absolute error as a percentage of the benchmark flow. | $\dfrac{100}{n}\sum_{ij}\lvert \dfrac{e_{ij}}{F^{bench}_{ij}} \rvert$ |
| Pearson's $r$ | Linear correlation; shows whether large benchmark flows also tend to be large adjusted flows. | $\mathrm{cor}(F^{adj}_{ij}, F^{bench}_{ij})$ |
| Spearman's $\rho$ | Rank correlation; shows whether the ordering of flows is similar. | $\mathrm{cor}(\mathrm{rank}(F^{adj}_{ij}), \mathrm{rank}(F^{bench}_{ij}))$ |
| Kullback-Leibler divergence | \[Level 4\] Directional difference between destination-share distributions. Lower values mean the destinations are allocated more similar to the reference flow. | $\sum_i\sum_j p^{ref}_{j\mid i}\log\dfrac{p^{ref}_{j\mid i}}{p^{comp}_{j\mid i}}$ |
| Jensen-Shannon divergence | \[Level 4\] Symmetric version of KL divergence. Lower values mean closer allocation. | $\dfrac{1}{2} KL(p^{ref}\mid\mid m) + \dfrac{1}{2} KL(p^{comp}\mid\mid m)$ |
| Residual-versus-benchmark-flow correlation | \[Level 5\] Association between remaining residuals and benchmark flow size. Values near zero suggest less obvious scale dependence. | $\mathrm{cor}(e_{ij}, F^{bench}_{ij})$ |
| Moran's $I$ | \[Level 5\] Spatial autocorrelation in area-level residuals using user-supplied neighbour links $w_{ik}$. Positive values suggest neighbouring areas have similar residuals. | $\dfrac{n}{W}\dfrac{\sum_i\sum_k w_{ik}(\bar e_i-\bar e)(\bar e_k-\bar e)}{\sum_i(\bar e_i-\bar e)^2}$ |
| Local Moran's $I$ and LISA class | \[Level 5\] Local spatial association for each area, with permutation pseudo $p$-values and cluster classes such as high-high, low-low, high-low and low-high. | $\dfrac{(\bar e_i-\bar e)\sum_k w_{ik}(\bar e_k-\bar e)}{\sum_i(\bar e_i-\bar e)^2/n}$ |
| Residual-versus-covariate correlation | \[Level 5\] Association between area-level residuals and a selected area covariate $x_i$. Values near zero suggest less obvious covariate-linked error. | $\mathrm{cor}(\bar e_i, x_i)$ |
:::

## Getting set up

::: key-point-box
**Important!** We illustrate how to implement different validation
checks for the results of data adjustment methods, across the different
levels described above. For level 1, we illustrate a broad selection of
data adjustment methods, which include `inverse_penetration`,
`selection_rate2`, `coefficient`, `bayes_gravity`,
`bayes_gravity_education`, `bayes_origin_pool` and the `unadjusted`
data. These are described in more detail in the [adjustment methods section](#sec-list-adjustment-methods). For simplicity, validation for levels 2-5 is
illustrated for a narrower selection of adjustment methods, namely
`inverse_penetration`, `coefficient`, `bayes_gravity_education` and
`unadjusted`.
:::

We use `debiasR` for adjustment and validation, `debiasRdata` for the
empirical travel-to-work data and `dplyr` and `ggplot2` for data
handling and plots.

```{r load-packages}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

```{r load-validation-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_validation_helpers.R",
  "_validation_helpers.R",
  "../vignettes/_validation_helpers.R",
  "../../vignettes/_validation_helpers.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_validation_helpers.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

***Prepare the validation data***. We use LAD travel-to-work data
throughout the vignette. The example combines raw mobile phone-derived
(MPD) flows with Census benchmark flows and the area information needed
by the adjustment methods. All methods are evaluated using the same LADs
and origin-destination pairs.

```{r load-validation-data}
example_data <- debiasR_example_data(
  n_areas = Inf,
  complete_grid = TRUE,
  geography = "lad"
)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage_validation <- example_data$coverage
covariates_validation <- example_data$covariates
distance_validation <- example_data$distance

area_names <- covariates_validation |>
  dplyr::select(area, name)
```

The main objects are: `mpd_od`, the raw MPD flows to be adjusted;
`benchmark_od`, the Census flows used as the validation benchmark;
`coverage_validation`, population and active-user coverage by area;
`covariates_validation`, area characteristics used by some adjustment
methods; `distance_validation`, distances between origins and
destinations. The resulting data cover `r example_data$metadata$n_areas`
LADs and `r example_data$metadata$n_mpd_od` origin-destination pairs.

***Create the results to validate.*** We first apply three deterministic
(non-Bayesian) adjustment methods that will be used in the validation
examples below. The function `adjust_all_methods()` applies the selected
methods to the same input data and returns a separate OD table for each
method. We also retain the original MPD flows as an unadjusted baseline.
This allows us to compare each adjusted result with both the Census
benchmark and the data before adjustment.

```{r fit-deterministic-methods}
deterministic_methods <- c(
  "inverse_penetration",
  "selection_rate2",
  "coefficient"
)

deterministic_results <- adjust_all_methods(
  mpd_od_df = mpd_od,
  coverage_df = coverage_validation,
  benchmark_od_df = benchmark_od,
  covariates_df = covariates_validation,
  covariate_col = "per_level4",
  methods = deterministic_methods
)

deterministic_results <- c(
  list(unadjusted = dplyr::mutate(mpd_od, flow_adj = flow)),
  deterministic_results
)

deterministic_method_results <- deterministic_results |>
  validate_flow_prepare_outputs(mpd_df = mpd_od)
```

Bayesian models take considerably longer to fit than the deterministic
methods, so we do not fit them while rendering this vignette. Instead,
`debiasR` provides Bayesian adjustment results fitted in advance using
the same LAD data. The specifications used to produce these results are
described in the [adjustment methods section](#sec-list-adjustment-methods) down below. The
following code chunk loads the saved results from the installed package
and separates them by adjustment method.

```{r load-bayesian-validation-rows}
bayesian_file <- system.file(
  "extdata",
  "v07-validation-bayesian-adjusted.csv",
  package = "debiasR"
)

bayesian_results <- read.csv(bayesian_file)

bayesian_method_results <- split(
  bayesian_results,
  bayesian_results$method
)
```

Then, bring together the results for all the adjustment methods
described in the [adjustment methods section](#sec-list-adjustment-methods), including those from
the deterministic methods obtained via `adjust_all_methods()` and from
the Bayesian methods which we precomputed and just loaded via the code
chunk above.

```{r assemble-validation-method-set}
all_method_results <- c(deterministic_method_results, bayesian_method_results)
```

## Validation measures

Next we illustrate how to implement different validation checks for the
data adjustment methods. We do so across the [different levels described
above](#sec-overview-levels).

### Level 1: Overall validation metrics

Level 1 compares complete OD tables. We compare the three deterministic
methods fitted above, three precomputed Bayesian models and the
unadjusted MPD flows.

```{r}
level1_methods <- c(
  "unadjusted",
  deterministic_methods,
  "bayes_gravity",
  "bayes_gravity_education",
  "bayes_origin_pool"
)

level1_method_results <- all_method_results[level1_methods]
```

<!-- Before calculating the validation measures, a hidden check confirms that -->

<!-- every method contains the same origin-destination pairs and the same -->

<!-- original MPD flows. This ensures that differences in the results reflect -->

<!-- the adjustment methods rather than differences in the data being -->

<!-- compared. -->

<!-- All methods are evaluated on the same `r nrow(mpd_od)` OD rows, -->

<!-- including zero-flow rows. MAPE is treated separately because percentage -->

<!-- errors are undefined when the benchmark flow is zero. -->

The `debiasR` function `validate_flow_overall()` calculates error and
correlation measures across all OD pairs. To apply it to each adjustment
method, this vignette uses the package helper
`validate_flow_overall_methods()`. Setting
`comparisons = "adjusted_vs_benchmark"` compares each result with the
Census benchmark.

```{r level1-overall-metrics}
overall_results <- validate_flow_overall_methods(
  adjusted_outputs = level1_method_results,
  benchmark_od_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
```

We first examine mean absolute error (MAE) and root mean squared error
(RMSE). Both are measured in flow counts and lower values indicate
closer agreement with the benchmark. The function
`validate_flow_overall()` also calculates other error measures,
including mean error, median absolute error and MAPE, as well as
correlation measures such as Pearson’s correlation and Spearman’s
correlation. The figure below uses `validate_flow_plot_metrics()` to
display MAE and RMSE. The other overall validation measures are retained
in `overall_results` and can be examined separately.

```{r level1-metric-plot}
validate_flow_plot_metrics(
  overall_results,
  error_measures = c("mae", "rmse"),
  comparisons = "adjusted_vs_benchmark",
  methods = level1_methods,
  method_labels = validation_plot_method_labels,
  sort = "ascending",
  sort_metric = "mae"
)
```

```{r level1-inference-values, include=FALSE}
level1_best_overall <- validate_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = level1_methods
)

level1_best_cor <- validate_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "pearson_r",
  methods = level1_methods,
  largest = TRUE
)
```

The methods are ordered from lowest to highest MAE. The unadjusted MPD
flows are kept at the bottom as a reference, regardless of their MAE. In
this example, `r level1_best_overall` has the lowest overall mean
absolute error against the benchmark. The strongest Pearson correlation
would be for `r level1_best_cor`. These measures can be related but one
is about distance from the benchmark, and the other is about similarity
in the overall pattern of flows.

::: key-point-box
**Remember!** The Bayesian models do not use benchmark OD flows when
they are fitted. Some deterministic methods do use benchmark
information, so this difference should be considered when interpreting
their validation results.
:::

### Level 2: Origin and destination marginals

Level 2 compares flows after aggregating them by geographical area. An
**origin marginal** is the total flow leaving an area, while a
**destination marginal** is the total flow entering an area. These
totals can reveal geographical differences that are hidden by the
system-wide measures used in Level 1.

::: key-point-box
**Remember!** From this level onwards, we use four results to keep the
detailed plots more readable. These are `inverse penetration`,
`coefficient adjustment`, `Bayesian gravity with education` and the
`unadjusted` MPD baseline.
:::

```{r select-validation-methods}
validation_methods <- c(
  "inverse_penetration",
  "coefficient",
  "bayes_gravity_education",
  "unadjusted"
)

validation_method_results <- all_method_results[validation_methods]
```

The package helper `validate_flow_margins()` sums the OD
flows by origin or destination. The package helper
`validate_flow_overall_methods()` then applies the `debiasR` function
`validate_flow_overall()` to each method.

```{r construct-level2-margins}
origin_margins <- validate_flow_margins(
  adjusted_outputs = validation_method_results,
  benchmark_od_df = benchmark_od,
  role = "origin"
)

origin_margin_metrics <- validate_flow_overall_methods(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_od_df = origin_margins$benchmark,
  comparisons = "adjusted_vs_benchmark"
)

destination_margins <- validate_flow_margins(
  adjusted_outputs = validation_method_results,
  benchmark_od_df = benchmark_od,
  role = "destination"
)

destination_margin_metrics <- validate_flow_overall_methods(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_od_df = destination_margins$benchmark,
  comparisons = "adjusted_vs_benchmark"
)
```

We produce a first figure to compare origin totals. The package helper
`validate_flow_residual_methods()` applies the function
`validate_flow_residuals()` to each method. The function
`validate_flow_plot_scatter()` plots the results. Each point represents one
origin. Points close to the dashed line have similar adjusted and
benchmark outflows.

```{r level2-origin-scatter, fig.width=9, fig.height=9}
origin_margin_residuals <- validate_flow_residual_methods(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_od_df = origin_margins$benchmark,
  top_n = 8
)

validate_flow_plot_scatter(
  origin_margin_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 2,
  point_size = 2.2,
  point_alpha = 0.8,
  sort = "ascending"
)
```

We produce a second figure to compare destination totals. Each point
represents one destination. Points close to the dashed line have similar
adjusted and benchmark inflows.

```{r level2-destination-scatter, fig.width=9, fig.height=9}
destination_margin_residuals <- validate_flow_residual_methods(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_od_df = destination_margins$benchmark,
  top_n = 8
)

validate_flow_plot_scatter(
  destination_margin_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 2,
  point_size = 2.2,
  point_alpha = 0.8,
  sort = "ascending"
)
```

```{r level2-inference-values, include=FALSE}
level2_origin_best <- validate_best_method(
  origin_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = validation_methods
)

level2_destination_best <- validate_best_method(
  destination_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = validation_methods
)
```

In this example, `r level2_origin_best` has the lowest MAE for origin
totals and `r level2_destination_best` has the lowest MAE for
destination totals. Generally, an adjustment method may perform
differently for the two margins because reproducing total outflows does
not necessarily mean that it also reproduces total inflows.

### Level 3: Individual OD-flow validation

Level 3 compares individual origin-destination pairs. This is the most
granular validation level and is the best place to inspect large local
errors. We continue with the four adjustment results selected in Level
2.

The function `validate_flow_residuals()` calculates signed and absolute
residuals for every OD pair. It also identifies the OD pairs with the
largest remaining errors. The package helper
`validate_flow_residual_methods()` applies this function to each adjustment
method.

```{r level3-residual-summary}
flow_residuals <- validate_flow_residual_methods(
  adjusted_outputs = validation_method_results,
  benchmark_od_df = benchmark_od,
  top_n = 12
)
```

We first use the function `validate_flow_plot_scatter()` to compare the
adjusted and benchmark flow for every OD pair. Each point represents one
origin-destination pair. Points on or near the dashed line have similar
adjusted and benchmark flows. Points farther from the line have larger
errors. For an adjusted-versus-benchmark comparison, the signed residual
is benchmark minus adjusted. A positive residual means that the adjusted
flow is below the benchmark, while a negative residual means that it is
above the benchmark. Point colour shows this direction and stronger
colours indicate larger differences.

```{r level3-flow-scatter, fig.width=9, fig.height=9}
validate_flow_plot_scatter(
  flow_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 2,
  point_alpha = 0.55,
  point_size = 1.2,
  sort = "ascending"
)
```

The exported `validate_flow_plot_residual_bands()` function provides a
complementary visualisation. It groups OD pairs according to the
absolute size of their residuals, measured in standard-deviation bands.
Larger shares in the lower residual bands indicate that more OD pairs
are close to the benchmark. Larger shares in the highest band indicate
that more OD pairs have severe remaining errors. The methods are ordered
by the share of OD pairs in the highest residual band, while the
unadjusted MPD flows remain at the bottom as a reference.

```{r level3-residual-bands}
validate_flow_plot_residual_bands(
  flow_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = validation_methods,
  method_labels = validation_plot_method_labels,
  band_method = "sd",
  sd_reference = "adjusted",
  orientation = "horizontal",
  label_min_share = 4,
  sort = "ascending",
  sort_metric = "severe_share"
)
```

### Level 4: Distributional allocation validation

Level 4 assesses whether flows from each origin are distributed across
destinations in similar proportions to the benchmark. For each origin,
the flows to all destinations are converted into shares that sum to one.
This separates the pattern of destination choice from the total amount
of flow. For example, an adjusted origin total could be too large while
still allocating the correct proportions to each destination.

The `debiasR` function `validate_flow_distribution()` compares these
origin-specific destination shares using Kullback-Leibler divergence and
Jensen-Shannon divergence. The Jensen-Shannon divergence provides a
symmetric measure of the difference between two distributions. Lower
values indicate that the adjusted and benchmark destination shares are
more similar. A value of zero would mean that they have identical
destination shares. The package helper
`validate_flow_distribution_methods()` applies this function to each
adjustment method.

```{r level4-distribution-summary}
distribution_results <- validate_flow_distribution_methods(
  adjusted_outputs = validation_method_results,
  benchmark_od_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
```

The function `validate_flow_plot_distribution_pairwise()` displays the
benchmark-weighted mean Jensen-Shannon divergence for each method. The
weighting gives more influence to origins with larger benchmark
outflows. The methods are ordered from lowest to highest divergence,
while the unadjusted MPD flows remain at the bottom as a reference.

```{r level4-distribution-pairwise, fig.width=9, fig.height=6}
validate_flow_plot_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = "adjusted_vs_benchmark",
  methods = validation_methods,
  method_labels = validation_plot_method_labels,
  plot_type = "comparison",
  sort = "ascending"
)
```

```{r level4-inference-values, include=FALSE}
distribution_summary_tbl <- dplyr::bind_rows(
  lapply(names(distribution_results), function(method_id) {
    out <- distribution_results[[method_id]]$summary
    out$method <- method_id
    out
  })
)

level4_best <- distribution_summary_tbl |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::slice_min(
    jsd_weighted_mean,
    n = 1,
    with_ties = FALSE
  ) |>
  dplyr::pull(method) |>
  validate_method_label()
```

To summarise the comparison, we identify the method with the lowest
benchmark-weighted Jensen-Shannon divergence. In this example,
`r level4_best` has the lowest benchmark-weighted Jensen-Shannon
divergence. This means that its destination shares are the closest to
the benchmark shares. It does not necessarily mean that the method has
the smallest errors in individual flow values or total flow magnitude.

### Level 5: Spatial and residual-structure diagnostics

Level 5 assesses wether the error measures between the adjusted flows
and the reference flows follow systematic patterns. A method may perform
well on average while still producing errors that are associated with
flow size, concentrated in neighbouring areas or related to area
characteristics. We propose three types of tests for residual structure:

-   correlation between residuals and benchmark flow size
-   spatial clustering among origin-level residuals
-   correlation between origin-level residuals and an area covariate

In this example, the area covariate is the percentage of residents with
Level 4 qualifications or above (`per_level4`). The spatial clustering checks require a definition of which areas are
neighbours. The vignette helper `validate_prepare_lad_lisa_inputs()`
constructs reproducible links between each LAD and its four nearest
neighbours using the centroid distances returned by
`debiasR_example_data()`. These are distance-based neighbours. The helper also looks for the LAD boundaries used by the map. It uses a
public ONS 2021 boundary file or a local boundary file supplied through
`DEBIAS_LAD_BOUNDARY_PATH`.

```{r prepare-residual-structure-inputs}
validation_lisa_inputs <- validate_prepare_lad_lisa_inputs(
  areas = covariates_validation$area,
  distance_df = distance_validation,
  distance_k = 4
)

structure_neighbors <- validation_lisa_inputs$neighbors
validation_lisa_boundaries <- validation_lisa_inputs$boundaries
validation_lisa_map_note <- validation_lisa_inputs$map_note

structure_covariates <- covariates_validation |>
  dplyr::select(area, per_level4)
```

```{r level5-neighbor-note, echo=FALSE}
cat(validation_lisa_inputs$note, "\n")
```

The `debiasR` function `validate_flow_residual_structure()` calculates
the residual-flow correlation, Moran’s I, Local Moran’s I and the
residual-covariate correlation. The package helper
`validate_flow_residual_structure_methods()` applies this function to each of the
four adjustment results.

```{r level5-residual-structure}
residual_structure <- validate_flow_residual_structure_methods(
  adjusted_outputs = validation_method_results,
  benchmark_od_df = benchmark_od,
  comparison = "adjusted_vs_benchmark",
  spatial_role = "origin",
  residual_aggregation = "mean",
  area_neighbors = structure_neighbors,
  local_moran = TRUE,
  covariate_df = structure_covariates,
  covariate_col = "per_level4",
  methods = validation_methods,
  local_moran_nsim = 199,
  local_moran_seed = 20260613
)
```

The function `validate_flow_plot_structure()` summarises the three level 5 tests proposed above. A residual-flow correlation close to zero indicates that
errors do not have an obvious linear relationship with benchmark flow
size. A residual-covariate correlation close to zero indicates less
evidence of a linear relationship with the selected area characteristic. For the spatial clustering test, we look at Moran’s $I$, which measures whether neighbouring areas have similar residuals. Positive values indicate spatial clustering, while values near zero indicate less evidence of an overall spatial pattern.

```{r level5-structure-plot}
validate_flow_plot_structure(
  residual_structure,
  methods = validation_methods,
  method_labels = validation_plot_method_labels
)
```

While global Moran’s $I$ summarises spatial clustering across all areas, the local
Moran’s $I$ examines each area separately and assigns it to a Local
Indicators of Spatial Association (LISA) class. For the adjusted-versus-benchmark comparison, positive residuals mean that adjusted flows are below the benchmark and negative residuals mean that they are above it. Therefore, a high-high area in a LISA map has a positive
residual and is surrounded by areas with positive residuals. A low-low
area has a negative residual and is surrounded by areas with negative
residuals. High-low and low-high areas are local spatial outliers. The function `validate_flow_plot_lisa_map()` maps these local patterns.
Areas without evidence of local spatial association at the selected
threshold are shown as not significant. The map uses raw permutation
pseudo $p$-values of 0.05, so it should be treated as an exploratory
screen rather than a multiple-testing-adjusted decision rule.

```{r level5-lisa-map, fig.width=10, fig.height=12}
if (is.null(validation_lisa_boundaries)) {
  cat(validation_lisa_map_note, "\n")
} else {
  validate_flow_plot_lisa_map(
    residual_structure,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = validation_methods,
    method_labels = validation_plot_method_labels,
    cluster_col = "lisa_quadrant",
    p_value_col = "p_value",
    p_value_threshold = 0.05,
    missing_fill = "#D1D5DB",
    boundary_colour = NA,
    boundary_linewidth = 0
  )
}
```



## Recommendation

Our recommendation is to adopt a sequence approach to validation. First,
we suggest to analyse the overall metrics to assess if the resulting
adjusted flows are broadly closer to the benchmark. Second, we recommend
evaluating origin and destination marginal diagnostics to determine the
precision of correction for area totals and identify systematic under-
or over-estimation for specific areas. Third, we suggest inspecting
individual origin-destinations pairs to evaluate corridors of special
interest, such as origin-destinations pairs that typically have large
amounts of interaction.

The validation methods presented here offer quantitative evidence to
assist in the selection of which method to use. However, this decision depends on a combination of qualitative and
quantitative factors:

-   *Data availability*. Adjustment methods require different levels of
    inputs. Simpler deterministic adjustment methods generally require
    fewer data inputs. Yet, they often require target benchmark flow
    data and these are precisely the data inputs that we often do not
    have. Benchmark flow data often come from censuses, which are
    considered the gold standard, but these data are generally collected
    every ten or five years and are outdated when a crisis occurs or a
    new evaluation of mobility is needed to support humanitarian
    operations, planning or policy decisions. Hence, using an adjustment
    method requiring census benchmark data would produce misleading
    results if it represented the mobility network at the time of the
    census rather than the mobility system during the period of
    interest. An adjustment method, such as the proposed Bayesian
    multilevel modelling approach, that is not reliant on such inputs
    may be more appropriate in such scenarios.
-   *Required measurement precision*. Adjustment methods have varying
    levels of precision in terms of scale reproducing population
    numbers. Prior work has focused on satisfying the
    requirement of proportionality by assessing the linear correlation
    between raw and/or adjusted mobile-phone-derived flows and benchmark
    flows. Yet, such assessment does not guarantee that raw or adjusted
    flows provide an accurate estimate in population numbers. Except for
    the selection rate II method, all the methods in `debiasR` seek to
    provide population numbers.
-   *Time constraint*. The adjustment methods in `debiasR` may demand more or less time. Such demand will depend on processes such as data gathering, data processing and calibrating adjustment methods. As suggested above, the use of digital trace data is often triggered by the rapid and urgent
    need to have "rough" population numbers to inform humanitarian,
    planning or policy decisions, to address a crisis or new
    intervention.

## Adjustment methods used in this vignette {#sec-list-adjustment-methods}

The table below states which adjustment methods are used in this vignette and the data inputs required by each of them. Not all methods learn from the same information. Yet, for illustration purposes, we still validate all outputs on the same benchmark OD table and the same OD rows.

```{r method-inputs, echo=FALSE}
validate_display_method_inputs(
  methods = unique(c(level1_methods, validation_methods))
)
```
