---
title: "Validation"
description: >
  Compare adjusted mobile-phone-derived origin-destination flows with benchmark
  flows across overall, marginal, OD-pair, distributional allocation, and
  residual-structure diagnostics.
vignette: >
  %\VignetteIndexEntry{Validation}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
format:
  html:
    html-math-method: mathml
    toc: true
execute:
  echo: true
  warning: false
  message: false
---

```{=html}
<style>
.template-quarto table {
  font-size: 14.5px;
}
.validation-table-small,
.validation-table-small table,
table.validation-table-small,
.template-quarto .validation-table-small {
  font-size: 13px;
}
.validation-table-compact,
.validation-table-compact table,
table.validation-table-compact,
.template-quarto .validation-table-compact {
  font-size: 13.5px;
}
</style>
```

## Aim

This vignette describes how to validate adjusted mobile-phone-derived
origin-destination (OD) flows against a benchmark OD table. The aim is simple:
after an adjustment method has been applied, we want to know whether the
adjusted flows are closer to the benchmark flows than the raw mobile-phone
flows were.

The key comparison is adjusted *versus* benchmark flows. This tells us how close
each output is to the target benchmark. The raw mobile-phone-derived table is
included as an unadjusted method, so the same adjusted-versus-benchmark
comparison also shows the starting point before correction.

A central requirement in this vignette is fairness. All methods are assessed
using the same validation data: the same raw OD table, the same benchmark OD
table, and the same full LAD origin-destination matrix. When a method needs
additional fitting inputs, such as coverage rates, distance or covariates,
those inputs are taken from the same prepared example data.
LAD means local authority district, a local government geography used in the
example data.

The example compares deterministic adjustment methods that can be fitted live
on the full LAD matrix during vignette rendering. These include simple coverage
weighting, benchmark-assisted calibration and benchmark-cell calibration. It
also includes one saved Bayesian `coverage_offset` result. Some methods use
benchmark totals or benchmark OD cells during fitting, so the validation result
should be read alongside the fitting inputs table rather than as a single
winner-takes-all ranking.

## Intuition

A practical validation strategy is to move from broad summaries to more local
diagnostics. This helps us avoid relying on one number. A method can look good
overall and still perform poorly for some origins, destinations or individual
OD pairs.

This vignette uses five levels.

**Level 1** summarises performance across the whole OD system. This tells us
whether a method is broadly close to the benchmark.

**Level 2** compares origin and destination totals. This checks whether adjusted
flows reproduce total outflows from each origin and total inflows to each
destination.

**Level 3** compares individual OD pairs. This shows where large local errors
remain after adjustment.

**Level 4** assesses whether each origin allocates its flows across destinations
in a similar way to the benchmark. This is about destination shares within
origins, not only total scale.

**Level 5** checks whether remaining residuals still show structure after
adjustment. For example, residuals may be larger for large benchmark flows,
clustered across neighbouring areas, or associated with an area covariate.

At each level, it is useful to separate error metrics from correlation metrics.
Error metrics tell us whether the numbers are close in size. Correlation
metrics tell us whether the broad pattern is similar. Both matter. A method can
have a high correlation with the benchmark and still be too high or too low in
absolute terms.

## Validation measures

Let $F^{adj}_{ij}$ be the adjusted flow from origin $i$ to destination
$j$, and let $F^{bench}_{ij}$ be the benchmark flow for the same OD pair.
Define the error as
$e_{ij} = F^{adj}_{ij} - F^{bench}_{ij}$. Positive errors mean the adjusted
flow is larger than the benchmark. Negative errors mean it is smaller. The
table uses adjusted-versus-benchmark notation.[^validation-measure-reuse]

::: {.validation-table-small}
| Metric | Description | Equation |
|:--|:--|:--|
| Mean error | Average signed error; shows whether flows are overestimated or underestimated on average. | $\frac{1}{n}\sum_{ij} e_{ij}$ |
| Mean absolute error | Average absolute error; shows the typical size of the error, ignoring sign. | $\frac{1}{n}\sum_{ij} \lvert e_{ij} \rvert$ |
| Root mean squared error | Error metric that gives more weight to large errors. | $\sqrt{\frac{1}{n}\sum_{ij} e_{ij}^{2}}$ |
| Median absolute error | Middle absolute error; less affected by a small number of very large errors. | $\mathrm{median}(\lvert e_{ij} \rvert)$ |
| Mean absolute percentage error | Average absolute error as a percentage of the benchmark flow. | $\frac{100}{n}\sum_{ij}\lvert \frac{e_{ij}}{F^{bench}_{ij}} \rvert$ |
| Pearson's r | Linear correlation; shows whether large benchmark flows also tend to be large adjusted flows. | $\mathrm{cor}(F^{adj}_{ij}, F^{bench}_{ij})$ |
| Spearman's rho | Rank correlation; shows whether the ordering of flows is similar. | $\mathrm{cor}(\mathrm{rank}(F^{adj}_{ij}), \mathrm{rank}(F^{bench}_{ij}))$ |
| Kullback-Leibler divergence | Level 4 metric. Directional difference between origin-conditioned destination-share distributions. Lower values mean the comparison flow allocates destinations more like the reference flow. | $\sum_i\sum_j p^{ref}_{j\mid i}\log\frac{p^{ref}_{j\mid i}}{p^{comp}_{j\mid i}}$ |
| Jensen-Shannon divergence | Level 4 metric. Symmetric companion to KL divergence for the same origin-conditioned destination-share distributions. Lower values mean closer allocation. | $\frac{1}{2} KL(p^{ref}\mid\mid m) + \frac{1}{2} KL(p^{comp}\mid\mid m)$ |
| Residual-versus-benchmark-flow correlation | Level 5 diagnostic. Association between remaining residuals and benchmark flow size. Values near zero suggest less obvious scale dependence. | $\mathrm{cor}(e_{ij}, F^{bench}_{ij})$ |
| Moran's I | Level 5 diagnostic. Spatial autocorrelation in area-level residuals using user-supplied neighbour links $w_{ik}$. Positive values suggest neighbouring areas have similar residuals. | $\frac{n}{W}\frac{\sum_i\sum_k w_{ik}(\bar e_i-\bar e)(\bar e_k-\bar e)}{\sum_i(\bar e_i-\bar e)^2}$ |
| Local Moran's I and LISA class | Level 5 diagnostic. Local spatial association for each area, with permutation pseudo p-values and cluster classes such as high-high, low-low, high-low and low-high. | $\frac{(\bar e_i-\bar e)\sum_k w_{ik}(\bar e_k-\bar e)}{\sum_i(\bar e_i-\bar e)^2/n}$ |
| Residual-versus-covariate correlation | Level 5 diagnostic. Association between area-level residuals and a selected area covariate $x_i$. Values near zero suggest less obvious covariate-linked error. | $\mathrm{cor}(\bar e_i, x_i)$ |
:::

[^validation-measure-reuse]: The equations are written for
    adjusted-versus-benchmark flows. The same formulas can be reused for
    raw-versus-benchmark or raw-versus-adjusted comparisons by replacing
    $F^{adj}_{ij}$ and $F^{bench}_{ij}$ with the two flow columns being
    compared.

## Getting ready

We start by loading the packages used in this vignette.

```{r load-packages}
library(debiasR)
library(debiasRdata)
library(dplyr)
library(ggplot2)
```

```{r load-validation-helpers, include=FALSE}
helper_candidates <- c(
  "vignettes/_validation_helpers.R",
  "_validation_helpers.R",
  "../vignettes/_validation_helpers.R",
  "../../vignettes/_validation_helpers.R"
)
helper_file <- helper_candidates[file.exists(helper_candidates)][1]
if (is.na(helper_file)) {
  stop(
    "Could not find vignettes/_validation_helpers.R. ",
    "Set the working directory to the debiasR package root or vignettes folder."
  )
}
source(helper_file)
```

We use the same empirical LAD travel-to-work example introduced in the earlier
vignettes. `debiasR_example_data()` prepares an aligned full LAD
origin-destination matrix. It returns raw MPD flows, benchmark flows, coverage
rates, covariates and distances for the same LAD areas.

```{r load-validation-data}
example_data <- debiasR_example_data(
  n_areas = Inf,
  complete_grid = TRUE,
  geography = "lad"
)

mpd_od <- example_data$mpd_od
benchmark_od <- example_data$benchmark_od
coverage_validation <- example_data$coverage
covariates_validation <- example_data$covariates
distance_validation <- example_data$distance

area_names <- covariates_validation |>
  dplyr::select(area, name)
```

The validation data contain `r example_data$metadata$n_areas` LADs and
`r example_data$metadata$n_mpd_od` OD rows.

The comparison fits the five deterministic adjustment methods live. The raw
MPD baseline is included as an unadjusted method, so it can be compared on the
same rows and with the same validation functions.

```{r fit-deterministic-methods}
deterministic_methods <- c(
  "inverse_penetration",
  "selection_rate",
  "selection_rate2",
  "raking_ratio",
  "coefficient"
)

deterministic_results <- adjust_all_methods(
  mpd_od_df = mpd_od,
  coverage_df = coverage_validation,
  benchmark_od_df = benchmark_od,
  covariates_df = covariates_validation,
  covariate_col = "per_level4",
  methods = deterministic_methods
)

deterministic_results <- c(
  list(unadjusted = dplyr::mutate(mpd_od, flow_adj = flow)),
  deterministic_results
)

deterministic_method_results <- deterministic_results |>
  validation_standardize_adjusted_outputs(mpd_df = mpd_od)
```

`adjust_all_methods()` keeps the deterministic fits in one place. It uses the
same MPD flows, coverage table, benchmark OD table and covariates for each
method. The unadjusted row is simply the raw MPD flow copied into `flow_adj`;
it is the reference starting point before adjustment.

The Bayesian rows are read from saved v07 files in `inst/extdata/`. The
vignette does not refit Bayesian models or run MCMC during render. The main
comparison uses `coverage_offset` results because those are on a true-flow
scale. The saved `reduced_form` result is checked but kept out of the five-level
comparison because it is an MPD-scale counterfactual rather than a true-flow
estimate.

```{r load-bayesian-validation-rows, include=FALSE}
bayesian_metadata <- validation_load_v07_bayesian_metadata()
bayesian_display <- validation_load_v07_bayesian_display()
bayesian_selection <- validation_load_v07_bayesian_selection()
validation_assert_bayesian_fingerprint(
  metadata = bayesian_metadata,
  mpd_df = mpd_od,
  coverage_df = coverage_validation,
  covariates_df = covariates_validation,
  distance_df = distance_validation
)

bayesian_method_results <- validation_load_v07_bayesian_outputs()
validation_assert_v07_bayesian_bundle(
  adjusted_outputs = bayesian_method_results,
  display = bayesian_display,
  metadata = bayesian_metadata,
  selection = bayesian_selection
)

bayesian_method_results <- bayesian_method_results |>
  validation_standardize_adjusted_outputs(mpd_df = mpd_od)
```

The selected Bayesian row is the best-performing saved `coverage_offset`
variant by Level 1 mean absolute error. This gives the vignette one Bayesian
comparison row while keeping the full set of saved Bayesian outputs available
for the advanced Bayesian vignette.

```{r assemble-validation-method-set, include=FALSE}
all_method_results <- c(deterministic_method_results, bayesian_method_results)

bayesian_overall_results <- validation_overall_results(
  adjusted_outputs = all_method_results[validation_bayesian_methods],
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)

selected_bayesian_method <- validation_best_method_id(
  bayesian_overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = validation_bayesian_methods
)

v07_validation_methods <- c(
  "unadjusted",
  deterministic_methods,
  selected_bayesian_method
)

v07_method_results <- all_method_results[v07_validation_methods]
selected_bayesian_label <- validation_method_label(selected_bayesian_method)
```

For the five validation levels, the vignette therefore compares the unadjusted
raw MPD baseline, five deterministic adjustment methods and
`r selected_bayesian_label`. The deterministic methods cover coverage
weighting, benchmark-margin calibration and benchmark-cell calibration. The
Bayesian row does not use benchmark OD cells during fitting, so its validation
scores should be read alongside the model-inputs table at the end.

Before validating, we check that every method output uses the same OD pairs and
the same raw MPD flow column. This is the main comparability check in the
vignette.

```{r validation-row-audit, include=FALSE}
row_audit <- validation_audit_shared_validation_rows(
  adjusted_outputs = v07_method_results,
  mpd_df = mpd_od,
  benchmark_df = benchmark_od
)
validation_assert_row_audit(row_audit)
```

The hidden audit confirms that all methods have the same
`r nrow(mpd_od)` OD pairs, no missing or extra pairs, and the raw MPD flow
column matches the same input data. If this check failed, the vignette would
stop rather than report metrics that mix adjustment performance with data
construction differences.

The support check covers the `r nrow(mpd_od)` OD rows used for validation.
Overall metrics below keep zero-flow rows for MAE, RMSE and correlation so that
methods are compared on the same support. MAPE is interpreted separately
because percentage errors cannot use zero benchmark flows as denominators.

## Level 1: Overall validation metrics

Level 1 compares whole OD tables. The main adjusted-versus-benchmark comparison
uses the same method set used throughout the rest of the vignette: unadjusted
raw MPD, the five deterministic adjustment methods and the selected Bayesian
`coverage_offset` variant.

```{r level1-overall-metrics, include=FALSE}
overall_results <- validation_overall_results(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
```

Start with the error metrics. Lower mean absolute error and RMSE mean the
compared flows are closer in size. The hidden validation object also includes
signed mean error and MAPE, but the figure focuses on the count-scale error
metrics that are easiest to compare at this level.

Correlation metrics answer a related but different question. A high Pearson
correlation means the method keeps the same broad pattern of large and small
flows as the benchmark. A high Spearman correlation means the ranking of flows is
similar. A method can have a high correlation and still have large errors if it
gets the pattern right but the scale wrong.

```{r level1-metric-plot}
plot_validation_metrics(
  overall_results,
  error_measures = c("mae", "rmse"),
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  sort = "ascending",
  sort_metric = "mae"
)
```

The plot focuses on MAE and RMSE because they are on the flow-count scale.
MAPE is useful as a percentage-based check, but it is less helpful beside
count-scale errors because it behaves differently when benchmark flows are
small. The adjusted methods are sorted by MAE; the unadjusted raw MPD baseline
is kept at the bottom as the reference starting point.

```{r level1-inference-values, include=FALSE}
level1_best_overall <- validation_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = v07_validation_methods
)
level1_best_cor <- validation_best_method(
  overall_results,
  comparison = "adjusted_vs_benchmark",
  metric = "pearson_r",
  methods = v07_validation_methods,
  largest = TRUE
)
```

In this example, `r level1_best_overall` has the lowest overall mean absolute
error against the benchmark. `r level1_best_cor` has the strongest Pearson
correlation. These are related but not identical statements: one is about
distance from the benchmark, and the other is about similarity in the overall
pattern of flows.

## Level 2: Origin and destination marginals

Level 2 compares geographical area totals. An **origin marginal** adds together
all flows leaving each origin. A **destination marginal** adds together all
flows arriving in each destination. This section uses the same seven-method
comparison set as Level 1.

This level is useful because an adjustment can look reasonable overall while
still overestimating or underestimating the flows of particular areas.

```{r level2-origin-marginal-metrics, include=FALSE}
origin_margins <- validation_build_marginal_tables(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  role = "origin"
)

origin_margin_metrics <- validation_overall_results(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_df = origin_margins$benchmark,
  comparisons = "adjusted_vs_benchmark"
)
```

```{r level2-destination-marginal-metrics, include=FALSE}
destination_margins <- validation_build_marginal_tables(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  role = "destination"
)

destination_margin_metrics <- validation_overall_results(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_df = destination_margins$benchmark,
  comparisons = "adjusted_vs_benchmark"
)
```

The scatterplots below show the data behind the marginal metrics. Each point is
an area total. Points near the dashed line have similar values in the two
datasets being compared. The colour shows the signed difference, so strong
colours flag areas where the gap is larger.

```{r level2-origin-scatter, fig.width=9, fig.height=9}
origin_margin_residuals <- validation_residual_results(
  adjusted_outputs = origin_margins$adjusted_outputs,
  benchmark_df = origin_margins$benchmark,
  top_n = 8
)

plot_validation_scatter(
  origin_margin_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 3,
  point_size = 2.2,
  point_alpha = 0.8,
  sort = "ascending"
)
```

```{r level2-destination-scatter, fig.width=9, fig.height=9}
destination_margin_residuals <- validation_residual_results(
  adjusted_outputs = destination_margins$adjusted_outputs,
  benchmark_df = destination_margins$benchmark,
  top_n = 8
)

plot_validation_scatter(
  destination_margin_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 3,
  point_size = 2.2,
  point_alpha = 0.8,
  sort = "ascending"
)
```

```{r level2-inference-values, include=FALSE}
level2_origin_best <- validation_best_method(
  origin_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = v07_validation_methods
)
level2_destination_best <- validation_best_method(
  destination_margin_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = v07_validation_methods
)
```

In this example, `r level2_origin_best` is closest to the benchmark origin
totals by mean absolute error, while `r level2_destination_best` is closest to
the benchmark destination totals. If the best method differs between origins
and destinations, the choice depends on the purpose of the analysis. Commuting
outflows, service catchments and destination demand do not always place the
same weight on the two margins.

## Level 3: Individual OD-flow validation

Level 3 compares individual origin-destination pairs. This is the most granular
validation level and is the best place to inspect large local errors. This
section uses the same method set as the rest of the vignette, so the scatter and
residual-band review line up with the earlier levels.

```{r level3-residual-summary, include=FALSE}
flow_residuals <- validation_residual_results(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  top_n = 12
)
```

The next scatterplot compares individual OD pairs. Each point is one OD pair.
The point colour is the signed difference between the two flows in the panel.
For adjusted-versus-benchmark panels, this is benchmark minus adjusted flow.
This makes outliers easier to see: points far from the dashed line and with
strong colour deserve attention.

```{r level3-flow-scatter, fig.width=9, fig.height=9}
plot_validation_scatter(
  flow_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  facet_ncol = 3,
  point_alpha = 0.55,
  point_size = 1.2,
  sort = "ascending"
)
```

The residual-band plot summarises how much of the OD table falls into small or
large residual bands. The residual used here is benchmark minus adjusted flow.
Lower shares in the high-residual bands indicate fewer large remaining
OD-pair errors.

```{r level3-residual-bands}
plot_validation_residual_bands(
  flow_residuals,
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  band_method = "sd",
  sd_reference = "adjusted",
  orientation = "horizontal",
  label_min_share = 4,
  sort = "ascending",
  sort_metric = "severe_share"
)
```

```{r level3-inference-values, include=FALSE}
level3_metrics <- validation_overall_results(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
level3_best <- validation_best_method(
  level3_metrics,
  comparison = "adjusted_vs_benchmark",
  metric = "mae",
  methods = v07_validation_methods
)
```

At the individual OD-pair level, `r level3_best` has the lowest mean absolute
error in this example. The residual-band plot is useful because the average can
hide a small number of very large local errors.

## Level 4: Distributional allocation validation

The previous levels compare scale, residuals and outliers. Level 4 asks a
different question: whether each output sends each origin's flows to
destinations in similar proportions to the benchmark OD matrix.

This is an origin-conditioned allocation check. For each origin, flows are
converted into destination shares. The method is then assessed on whether those
shares resemble the benchmark shares. This matters because a method can improve
the total scale of flows while still allocating an origin's flows to the wrong
destinations.

```{r level4-distribution-summary, include=FALSE}
distribution_results <- validation_distribution_results(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  comparisons = "adjusted_vs_benchmark"
)
```

The comparison plot below uses weighted Jensen-Shannon divergence to show which
pairs of flow distributions are closest after converting flows into
origin-conditioned destination shares. Lower values mean the destination-share
distribution is closer to the reference distribution. The weighted version gives
more influence to origins with larger benchmark outflows.

```{r level4-distribution-pairwise, fig.width=9, fig.height=6}
plot_validation_distribution_pairwise(
  distribution_results,
  metric = "jsd",
  value = "weighted_mean",
  comparisons = "adjusted_vs_benchmark",
  methods = v07_validation_methods,
  method_labels = validation_plot_method_labels,
  plot_type = "comparison",
  sort = "ascending"
)
```

```{r level4-inference-values, include=FALSE}
distribution_summary_tbl <- dplyr::bind_rows(lapply(names(distribution_results), function(method_id) {
  out <- distribution_results[[method_id]]$summary
  out$method <- method_id
  out
}))
level4_best <- distribution_summary_tbl |>
  dplyr::filter(comparison == "adjusted_vs_benchmark") |>
  dplyr::slice_min(jsd_weighted_mean, n = 1, with_ties = FALSE) |>
  dplyr::pull(method) |>
  validation_method_label()
```

In this example, `r level4_best` has the lowest benchmark-weighted
Jensen-Shannon divergence in the adjusted-versus-benchmark comparison. This is
not the same as having the lowest OD-flow error. It means the method is closest
to the benchmark in terms of the destination shares sent from each origin.

## Level 5: Spatial and residual-structure diagnostics

The previous levels can show that an adjustment improves average error, area
totals, individual OD pairs and destination-share allocation. Level 5 asks
whether the remaining residuals still look systematic.

Here we check three simple structures. First, are residuals associated with the
size of benchmark flows? Second, are origin-level residuals spatially clustered
across neighbouring areas? Third, are origin-level residuals associated with a
known area covariate? In this example, the covariate is the percentage of
residents with Level 4 qualifications or above (`per_level4`).

This section uses the same seven-method comparison set as the earlier plots.
The optional Moran's I and LISA diagnostics need neighbour links. The Local
Moran statistic itself uses reproducible four-nearest-neighbour links derived
from the real LAD centroid distances returned by `debiasR_example_data()`.
The LISA map uses a simplified ONS 2021 LAD boundary file when `sf` and
network access are available. Readers can also set `DEBIAS_LAD_BOUNDARY_PATH`
to a local `sf`-readable boundary file. The diagnostic should be interpreted
as distance-neighbour rather than contiguity-neighbour LISA.

```{r prepare-residual-structure-inputs}
validation_lisa_inputs <- validation_prepare_lad_lisa_inputs(
  areas = covariates_validation$area,
  distance_df = distance_validation,
  distance_k = 4
)
structure_neighbors <- validation_lisa_inputs$neighbors
validation_lisa_boundaries <- validation_lisa_inputs$boundaries
validation_lisa_map_note <- validation_lisa_inputs$map_note

structure_covariates <- covariates_validation |>
  dplyr::select(area, per_level4)

structure_methods <- v07_validation_methods
```

```{r level5-neighbor-note, echo=FALSE}
cat(validation_lisa_inputs$note, "\n")
```

```{r level5-residual-structure, include=FALSE}
residual_structure <- validation_structure_results(
  adjusted_outputs = v07_method_results,
  benchmark_df = benchmark_od,
  area_neighbors = structure_neighbors,
  covariate_df = structure_covariates,
  covariate_col = "per_level4",
  methods = structure_methods,
  local_moran_nsim = 199,
  local_moran_seed = 20260613
)
```

Values close to zero are usually preferable for the residual-flow and
residual-covariate correlations, because they suggest less obvious residual
structure. For Moran's I, values close to zero suggest less spatial clustering
under the supplied neighbour definition. These are diagnostic checks, not final
proof that residuals are random.

```{r level5-structure-plot}
plot_validation_structure(
  residual_structure,
  methods = structure_methods,
  method_labels = validation_plot_method_labels
)
```

Local Moran's I provides a local version of the spatial clustering check. The
LISA class labels combine the sign of an area's centred residual with the
weighted residual pattern in its neighbours. The p-values are permutation
pseudo p-values from the supplied neighbour links, so they are best used as
exploratory flags for local review.

The LISA map below uses the same Local Moran output. Areas with raw permutation
pseudo p-values above 0.05 are shown as not significant. The underlying Local
Moran output records both raw and adjusted p-values, so the map should be read
as an exploratory local screen rather than a multiple-testing-adjusted decision
rule. Boundary polygons come from an ONS LAD boundary file, unless a local path
is supplied with `DEBIAS_LAD_BOUNDARY_PATH`. LADs that are present in the
boundary file but outside the validation support are retained in the same grey
used for "not significant" areas rather than dropped from the map. Scottish
boundary polygons are omitted because the validation example focuses on the
England and Wales support used by the LAD data.

```{r level5-lisa-map, fig.width=10, fig.height=12}
if (is.null(validation_lisa_boundaries)) {
  cat(validation_lisa_map_note, "\n")
} else {
  plot_validation_lisa_map(
    residual_structure,
    boundaries = validation_lisa_boundaries,
    boundary_area_col = "area",
    comparisons = "adjusted_vs_benchmark",
    methods = structure_methods,
    method_labels = validation_plot_method_labels,
    cluster_col = "lisa_quadrant",
    p_value_col = "p_value",
    p_value_threshold = 0.05,
    missing_fill = "#D1D5DB",
    boundary_colour = NA,
    boundary_linewidth = 0
  )
}
```

## Recommendation

Our recommendation is to adopt a sequence approach to validation. First, we
suggest to analyse the overall metrics to assess if the resulting adjusted flows
are broadly closer to the benchmark. Second, we recommend evaluating origin and
destination marginal diagnostics to determine the precision of correction for
area totals, and identify systematic under- or over-estimation for specific
areas. Third, we suggest inspecting individual origin-destinations pairs to
evaluate corridors of special interest, such as origin-destinations pairs that
typically have large amounts of interaction.

The validation methods presented here offer quantitative evidence to assist in
the selection of which method to use. However, this decision is not binary and
depends on a combination of qualitative and quantitative factors:

- *Data availability*. Adjustment methods require different levels of inputs.
  Simpler deterministic adjustment methods generally require fewer data inputs.
  Yet, they often require target benchmark flow data, and these are precisely
  the data inputs that we often do not have. Benchmark flow data often come
  from censuses, which are considered the gold standard, but these data are
  generally collected every ten or five years and are outdated when a crisis
  occurs or a new evaluation of mobility is needed to support humanitarian
  operations, planning or policy decisions. Hence, using an adjustment method
  requiring census benchmark data would produce misleading results if it
  represented the mobility network at the time of the census rather than the
  mobility system during the period of interest. An adjustment method, such as
  the proposed Bayesian multilevel modelling approach, that is not reliant on
  such inputs may be more appropriate in such scenarios.
- *Required measurement precision*. Adjustment methods have varying levels of
  precision in terms of scale reproducing population numbers. Historically,
  prior work has focused on satisfying the requirement of proportionality by
  assessing the linear correlation between raw and/or adjusted
  mobile-phone-derived flows and benchmark flows. Yet, such assessment does not
  guarantee that raw or adjusted flows provide an accurate estimate in
  population numbers. They often still represent number of devices, rather than
  people. Except for the selection rate II method, all the methods in `debiasR`
  seek to provide population numbers.
- *Time constraint*. The adjustment methods in `debiasR` may demand more or
  less time. Such demand will depend on processes such as data gathering, data
  processing and calibrating adjustment methods. These demands normally
  intersect and often come into conflict with the speed at which flow estimates
  are needed. As suggested above, the use of digital trace data is often
  triggered by the rapid and urgent need to have "rough" population numbers to
  inform humanitarian, planning or policy decisions, to address a crisis or new
  intervention.

## Model inputs used in this vignette

The table below states which inputs each method uses. This matters because
methods do not all learn from the same information. We still validate all
outputs on the same benchmark OD table and the same OD rows.

```{r method-inputs, echo=FALSE}
validation_display_method_inputs(
  methods = v07_validation_methods
)
```
